---
title: "ggfunction"
output:
  github_document:
    toc: false
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/readme-",
  out.width = "60%",
  fig.width = 6,
  fig.height = 4,
  warning = FALSE,
  message = FALSE
)
library(ggfunction)
library(ggplot2)
```

<!-- badges: start -->
<!-- badges: end -->

**ggfunction** extends [**ggplot2**](https://ggplot2.tidyverse.org/) with geoms and stats for plotting mathematical functions. It also provides more specialized geoms/stats for plotting *statistical* functions related to probability distributions.


## Overview

The package is organized around two families of geoms:

| Family | Geom | Maps | Description |
|--------|------|------|-------------|
| **Dimensional** | `geom_function_1d_1d()` | $\mathbb{R} \to \mathbb{R}$ | Scalar functions with optional interval shading |
| | `geom_function_1d_2d()` | $\mathbb{R} \to \mathbb{R}^2$ | Parametric curves |
| | `geom_function_2d_1d()` | $\mathbb{R}^2 \to \mathbb{R}$ | Scalar fields (raster, contour, filled contour) |
| | `geom_function_2d_2d()` | $\mathbb{R}^2 \to \mathbb{R}^2$ | Vector field streamlines |
| **Probability** | `geom_pdf()` | | Probability density function |
| | `geom_cdf()` | | Cumulative distribution function |
| | `geom_pmf()` | | Probability mass function (lollipop) |
| | `geom_qf()` | | Quantile function |
| | `geom_discrete_cdf()` | | Discrete CDF (step function) |
| | `geom_survival()` | | Survival function $S(x) = 1 - F(x)$ |
| | `geom_hf()` | | Hazard function $h(x) = f(x)/S(x)$ |

## Dimensional Taxonomy

### Scalar functions ($\mathbb{R} \to \mathbb{R}$)

At a basic level, `geom_function_1d_1d()`{.R} is essentially a re-implementation of `ggplot2::geom_function()`:

```{r 1d-1d}
ggplot() +
  geom_function_1d_1d(fun = sin, xlim = c(0, 2*pi))
```

However, `geom_function_1d_1d()`{.R} also allows for shading to the $x$ axis:

```{r 1d-1d-shade}
ggplot() +
  geom_function_1d_1d(fun = sin, xlim = c(-3, 3), shade_from = -1, shade_to = 1)
```

### Parametric curves ($\mathbb{R} \to \mathbb{R}^2$)

The graph of a function $\phi(t) = (x(t), y(t))$ is a curve in the plane $\mathbb{R}^{2}$.

```{r 1d-2d}
f <- function(t) c(sin(t), t * cos(t))
ggplot() +
  geom_function_1d_2d(fun = f, T = 20)
```

Tail points can be added with `tail_point = TRUE`:

```{r 1d-2d-tail}
ggplot() +
  geom_function_1d_2d(fun = f, T = 20, tail_point = TRUE)
```

### Scalar fields ($\mathbb{R}^2 \to \mathbb{R}$)

Visualize a scalar field as a raster, contour lines, or filled contours using the `type` argument.

```{r 2d-1d-raster}
f <- function(v) exp(-(v[1]^2 + v[2]^2) / 2)

ggplot() +
  geom_function_2d_1d(fun = f, xlim = c(-3, 3), ylim = c(-3, 3))
```

```{r 2d-1d-contour}
ggplot() +
  geom_function_2d_1d(fun = f, xlim = c(-3, 3), ylim = c(-3, 3), type = "contour")
```

```{r 2d-1d-filled}
ggplot() +
  geom_function_2d_1d(fun = f, xlim = c(-3, 3), ylim = c(-3, 3), type = "contour_filled")
```

### Vector fields ($\mathbb{R}^2 \to \mathbb{R}^2$)

Draw streamlines from a vector field.

```{r 2d-2d}
f <- function(u) c(-u[2], u[1])
ggplot() +
  geom_function_2d_2d(fun = f, xlim = c(-1, 1), ylim = c(-1, 1))
```

## Probability Distributions

### PDF

Shade below a quantile, between two quantiles, or shade the tails.

```{r geom-pdf}
ggplot() +
  geom_pdf(fun = dnorm, xlim = c(-3, 3))
```

```{r geom-pdf-p}
ggplot() +
  geom_pdf(fun = dnorm, xlim = c(-3, 3), p = 0.975)
```

```{r pdf-twosided}
# Middle 95%
ggplot() +
  geom_pdf(fun = dnorm, xlim = c(-3, 3),
    p_lower = 0.025, p_upper = 0.975)
```

```{r pdf-tails}
# Both tails
ggplot() +
  geom_pdf(fun = dnorm, xlim = c(-3, 3),
    p_lower = 0.025, p_upper = 0.975, shade_outside = TRUE)
```

### CDF

```{r geom-cdf}
ggplot() +
  geom_cdf(fun = pnorm, xlim = c(-3, 3))
```

```{r geom-cdf-p}
ggplot() +
  geom_cdf(fun = pnorm, xlim = c(-3, 3), p = 0.975)
```

### PMF (lollipop)

```{r geom-pmf}
ggplot() +
  geom_pmf(fun = dbinom, args = list(size = 10, prob = 0.3), xlim = c(0, 10))
```

### Quantile function

```{r geom-qf}
ggplot() +
  geom_qf(fun = qnorm, args = list(mean = 0, sd = 1))
```

### Discrete CDF (step function)

```{r discrete-cdf}
ggplot() +
  geom_discrete_cdf(fun = dbinom, args = list(size = 10, prob = 0.5), xlim = c(0, 10))
```

### Survival function

$S(x) = 1 - F(x)$

```{r survival}
ggplot() +
  geom_survival(fun = pexp, args = list(rate = 0.5), xlim = c(0, 10))
```

### Hazard function

$h(x) = f(x) / S(x)$

```{r hazard}
ggplot() +
  geom_hf(pdf_fun = dexp, cdf_fun = pexp, args = list(rate = 0.5), xlim = c(0.01, 10))
```

## Installation

```{r installation, eval=FALSE}
# install.packages("pak")
pak::pak("dusty-turner/ggfunction")
```

## Getting help

- Browse the [vignette](vignettes/ggfunction.Rmd) for a guided walkthrough
- File bugs or feature requests at <https://github.com/dusty-turner/ggfunction/issues>
