---
title: "ggfunction"
output:
  github_document:
    toc: false
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/readme-",
  out.width = "60%",
  fig.width = 6,
  fig.height = 4,
  warning = FALSE,
  message = FALSE
)
```

<!-- badges: start -->
[![Project Status: Active - The project has reached a stable, usable state and is being actively developed.](https://www.repostatus.org/badges/latest/active.svg)](https://www.repostatus.org/#active)
[![Lifecycle: experimental](https://img.shields.io/badge/lifecycle-experimental-orange.svg)](https://lifecycle.r-lib.org/articles/stages.html#experimental)
[![License: MIT](https://img.shields.io/badge/license-MIT-blue.svg)](https://opensource.org/licenses/MIT)
<!-- badges: end -->

**ggfunction** extends [**ggplot2**](https://ggplot2.tidyverse.org/) with geoms and stats for plotting mathematical functions and probability distributions directly from function objects. Supply a function and a domain; **ggfunction** handles evaluation, rendering, and shading.



## Overview

The package is organized around two families of geoms:

| Family | Geom | Maps | Description |
|--------|------|------|-------------|
| **Dimensional** | `geom_function_1d_1d()` | $\mathbb{R} \to \mathbb{R}$ | Scalar functions with optional interval shading |
| | `geom_function_1d_2d()` | $\mathbb{R} \to \mathbb{R}^2$ | Parametric curves |
| | `geom_function_2d_1d()` | $\mathbb{R}^2 \to \mathbb{R}$ | Scalar fields (raster, contour, filled contour) |
| | `geom_function_2d_2d()` | $\mathbb{R}^2 \to \mathbb{R}^2$ | Vector field streamlines |
| **Probability** | `geom_pdf()` | | Probability density function |
| | `geom_pmf()` | | Probability mass function (lollipop) |
| | `geom_cdf()` | | Cumulative distribution function |
| | `geom_cdf_discrete()` | | Discrete CDF (step function) |
| | `geom_survival()` | | Survival function $S(x) = 1 - F(x)$ |
| | `geom_survival_discrete()` | | Discrete survival function (step function) |
| | `geom_qf()` | | Quantile function |
| | `geom_qf_discrete()` | | Discrete quantile function (step function) |
| | `geom_hf()` | | Hazard function $h(x) = f(x)/S(x)$ |




## Dimensional Taxonomy

### Scalar functions: `geom_function_1d_1d()`

`geom_function_1d_1d()` evaluates a function $f\colon \mathbb{R} \to \mathbb{R}$ at `n` equally-spaced points over `xlim` and draws it as a curve, extending `ggplot2::stat_function()` with built-in region shading. The following example plots the sine function over one full period.

```{r 1d-1d}
library("ggfunction")

ggplot() +
  geom_function_1d_1d(fun = sin, xlim = c(0, 2*pi))
```

**Shading intervals.** The `shade_from` and `shade_to` parameters fill the region between the curve and the $x$-axis over a specified interval. Here we shade the standard normal density between $-1$ and $1$, corresponding to approximately 68% of the total area.

```{r 1d-1d-shade}
ggplot() +
  geom_function_1d_1d(fun = sin, xlim = c(-3, 3), shade_from = -1, shade_to = 1)
```






### Parametric curves: `geom_function_1d_2d()`

`geom_function_1d_2d()` evaluates a parametric curve $\boldsymbol{\gamma}(t) = (x(t),\, y(t))$ over the range `tlim` with step size `dt`, coloring the path by the time parameter by default. The following example traces the lemniscate of Bernoulli -- a figure-eight curve along which the product of distances to the two foci is constant -- $\boldsymbol{\gamma}(t) = \bigl(\cos t\,/\,(1 + \sin^2 t),\; \sin t\cos t\,/\,(1 + \sin^2 t)\bigr)$.

```{r 1d-2d}
lemniscate <- function(t) c(cos(t) / (1 + sin(t)^2), sin(t) * cos(t) / (1 + sin(t)^2))

ggplot() +
  geom_function_1d_2d(fun = lemniscate, tlim = c(0, 1.9 * pi))
```

**Arrowheads and tail points.** Setting `tail_point = TRUE` and passing `grid::arrow()` to `arrow` marks the starting position of the curve with a point and the tail with an arrowhead, respectively, useful for showing an initial condition and direction of the curve.

```{r 1d-2d-arrows-and-tail}
ggplot() +
  geom_function_1d_2d(
    fun = lemniscate, tlim = c(0, 1.9 * pi), tail_point = TRUE,
    arrow = grid::arrow(angle = 30, length = grid::unit(0.02, "npc"), type = "closed")
  )
```

**Parameterized families.** Extra parameters for `fun` are passed via `args`. The following example traces a Lissajous figure, a closed curve that results when two sinusoids with a rational frequency ratio are combined.

```{r 1d-2d-lissajous}
lissajous <- function(t, a = 3, b = 2, delta = pi/2) {
  c(sin(a * t + delta), sin(b * t))
}

ggplot() +
  geom_function_1d_2d(
    fun = lissajous, tlim = c(0, 1.9 * pi),
    args = list(a = 3, b = 2, delta = pi/2)
  )
```





### Scalar fields: `geom_function_2d_1d()`

`geom_function_2d_1d()` evaluates a scalar field $f\colon \mathbb{R}^2 \to \mathbb{R}$ on an $n \times n$ grid over `xlim` $\times$ `ylim`. The function must accept a numeric vector of length 2 and return a scalar; extracting components with `x <- v[1]; y <- v[2]` is the recommended pattern. The example below shows a Gaussian bump $f(x,y) = \exp\!\bigl(-(x^2+y^2)/2\bigr)$ rendered as a raster heatmap (the default `type`).

```{r 2d-1d-raster}
f <- function(v) {
  x <- v[1]; y <- v[2]
  sin(2*pi*x) + sin(2*pi*y)
}

ggplot() +
  geom_function_2d_1d(fun = f, xlim = c(-1, 1), ylim = c(-1, 1))
```

**Contour modes.** The `type` argument switches among three visual encodings of the same field. `"contour"` draws iso-level curves colored by level value; `"contour_filled"` draws filled regions between levels; both respond to `bins`, `binwidth`, and `breaks`.

```{r 2d-1d-contour}
ggplot() +
  geom_function_2d_1d(fun = f, xlim = c(-1, 1), ylim = c(-1, 1), type = "contour")
```

```{r 2d-1d-filled}
ggplot() +
  geom_function_2d_1d(fun = f, xlim = c(-1, 1), ylim = c(-1, 1), type = "contour_filled")
```





### Vector fields: `geom_function_2d_2d()`

`geom_function_2d_2d()` visualizes a vector field $\mathbf{F}\colon \mathbb{R}^2 \to \mathbb{R}^2$ via [**ggvfields**](https://github.com/dusty-turner/ggvfields). The function should accept a length-2 vector and return a length-2 vector. By default (`type = "vector"`) it draws short arrows at each grid point colored by field magnitude; `type = "stream"` switches to integral-curve streamlines. The following example shows the rotation field $\mathbf{F}(x,y) = (-y,\, x)$.

```{r 2d-2d}
f <- function(u) {
  x <- u[1]; y <- u[2]
  c(-y, x)
}

ggplot() +
  geom_function_2d_2d(fun = f, xlim = c(-1, 1), ylim = c(-1, 1))
```

**Stream field.** Setting `type = "stream"` renders the field as integral curves computed by numerical integration, colored by average speed along each curve.

```{r 2d-2d-stream}
ggplot() +
  geom_function_2d_2d(fun = f, xlim = c(-1, 1), ylim = c(-1, 1), type = "stream")
```









## Probability Distributions

The probability family provides a geom for each of the standard functions associated with a distribution. Each accepts a plain R function (e.g. `dnorm`, `pnorm`) and `xlim`; distribution parameters are passed via `args`.




### PDF: `geom_pdf()`

`geom_pdf()` draws a probability density function as a filled area. It validates that the supplied function integrates to 1 over `xlim`, warning via [**cli**](https://cli.r-lib.org/) if it does not. The basic call draws the full density with no shading.

```{r geom-pdf}
ggplot() +
  geom_pdf(fun = dnorm, xlim = c(-3, 3))
```

**Single threshold.** The `p` parameter shades from the left boundary up to the $p$-quantile (`lower.tail = TRUE`, the default). Setting `lower.tail = FALSE` shades the upper tail instead. Here we shade the lower 97.5%.

```{r geom-pdf-p}
ggplot() +
  geom_pdf(fun = dnorm, xlim = c(-3, 3), p = 0.975)
```

**Two-sided interval.** `p_lower` and `p_upper` together shade the central region between two quantiles---the natural picture for a confidence interval. The following example shades the central 95%.

```{r pdf-twosided}
ggplot() +
  geom_pdf(fun = dnorm, xlim = c(-3, 3), p_lower = 0.025, p_upper = 0.975)
```

**Tail shading.** Setting `shade_outside = TRUE` inverts the two-sided region, shading both tails. This is the rejection region of a two-sided hypothesis test at level $\alpha = 0.05$.

```{r pdf-tails}
ggplot() +
  geom_pdf(
    fun = dnorm, xlim = c(-3, 3),
    p_lower = 0.025, p_upper = 0.975, shade_outside = TRUE
  )
```

**Highest density region.** `shade_hdr` shades the smallest region of the domain containing the specified probability mass---the [highest density region](https://en.wikipedia.org/wiki/Credible_interval#Highest_density_interval) (HDR). For multimodal densities this region can be disconnected. The following example uses an asymmetric mixture of two normals; the 80% HDR captures both modes as two disjoint intervals, with more area allocated to the taller, narrower component.

```{r pdf-hdr}
f_mix <- function(x) 0.6 * dnorm(x, mean = -2, sd = 0.6) + 0.4 * dnorm(x, mean = 2, sd = 1.2)
ggplot() +
  geom_pdf(fun = f_mix, xlim = c(-5, 6), shade_hdr = 0.8)
```


### PMF: `geom_pmf()`

`geom_pmf()` evaluates a probability mass function at each integer in `xlim` and renders the result as a lollipop chart---vertical segments capped with points. Distribution parameters are supplied via `args`. The following example plots a $\text{Binomial}(10, 0.3)$ distribution.

```{r geom-pmf}
ggplot() +
  geom_pmf(fun = dbinom, xlim = c(0, 10), args = list(size = 10, prob = 0.3))
```

**Single threshold.** The `p` parameter shades lollipops up to the $p$-quantile (grey dashed sticks mark the unshaded region). Here we shade the lower 80% of a $\text{Binomial}(10, 0.5)$ distribution.

```{r geom-pmf-p}
ggplot() +
  geom_pmf(fun = dbinom, xlim = c(0, 10), args = list(size = 10, prob = 0.5), p = 0.8)
```

**Highest density region.** `shade_hdr` shades the smallest set of support points whose total probability mass meets or exceeds the target coverage. For a symmetric unimodal PMF this is a central interval; for a skewed distribution it is asymmetric. Here the 70% HDR of a $\text{Binomial}(10, 0.3)$ distribution is highlighted.

```{r geom-pmf-hdr}
ggplot() +
  geom_pmf(fun = dbinom, xlim = c(0, 10), args = list(size = 10, prob = 0.3),
    shade_hdr = .70)
```

```{r geom-pmf-hdr-2}
ggplot() +
  geom_pmf(fun = dbinom, xlim = c(0, 10), args = list(size = 10, prob = 0.3),
    shade_hdr = .80)
```

**Explicit support.** When the support is not a sequence of consecutive integers, pass the exact support points via `support`. Here we plot the distribution of the sample mean $\bar X$ of 10 iid $\text{Bernoulli}(0.3)$ draws. The support is $\{0, 0.1, 0.2, \ldots, 1\}$ and the PMF is binomial: $P(\bar X = k/10) = \binom{10}{k}0.3^k 0.7^{10-k}$.

```{r geom-pmf-support}
f_mean <- function(x, prob) dbinom(round(x * 10), size = 10, prob = prob)
ggplot() +
  geom_pmf(fun = f_mean, support = seq(0, 1, by = 0.1), args = list(prob = 0.3))
```


### CDF: `geom_cdf()` and `geom_cdf_discrete()`

The cumulative distribution function $F(x) = P(X \le x)$ is available for both continuous and discrete distributions.

**Continuous (`geom_cdf()`).** Draws the CDF as a line. Here we plot the standard normal CDF.

```{r geom-cdf}
ggplot() +
  geom_cdf(fun = pnorm, xlim = c(-3, 3))
```

**Discrete (`geom_cdf_discrete()`).** Takes a PMF, accumulates it into the right-continuous step-function CDF, and renders it with horizontal segments, dashed vertical jumps, open circles at the lower limit of each jump, and closed circles at the achieved value. The following example plots the $\text{Binomial}(10, 0.5)$ CDF.

```{r discrete-cdf}
ggplot() +
  geom_cdf_discrete(fun = dbinom, xlim = c(0, 10), args = list(size = 10, prob = 0.5))
```

**Parameterized families.** Here we plot a $\text{Poisson}(5)$ discrete CDF.

```{r discrete-cdf-pois}
ggplot() +
  geom_cdf_discrete(fun = dpois, xlim = c(0, 15), args = list(lambda = 5))
```

**Hiding points and lines.** `show_points = FALSE` and `show_vert = FALSE` remove the endpoint circles and vertical jump segments, leaving only the horizontal staircase.

```{r discrete-cdf-no-points-lines}
ggplot() +
  geom_cdf_discrete(fun = dbinom, xlim = c(0, 10), args = list(size = 10, prob = 0.5),
    show_points = FALSE, show_vert = FALSE)
```

**Explicit support.** Pass non-integer support points directly via `support`. Here we plot the CDF of the sample mean of 10 iid $\text{Bernoulli}(0.3)$ draws.

```{r discrete-cdf-support}
f_mean <- function(x, prob) dbinom(round(x * 10), size = 10, prob = prob)
ggplot() +
  geom_cdf_discrete(fun = f_mean, support = seq(0, 1, by = 0.1), args = list(prob = 0.3))
```


### Survival function: `geom_survival()` and `geom_survival_discrete()`

The survival function $S(x) = 1 - F(x) = P(X > x)$ gives the probability of surviving past $x$.

**Continuous (`geom_survival()`).** Accepts a CDF via `fun`. The following example shows the survival function of an $\text{Exponential}(0.5)$ distribution, which decays as $S(x) = e^{-0.5x}$.

```{r survival}
ggplot() +
  geom_survival(fun = pexp, xlim = c(0, 10), args = list(rate = 0.5))
```

**Discrete (`geom_survival_discrete()`).** Takes a PMF and renders $S(x) = 1 - F(x)$ as a right-continuous step function using the same visual conventions as `geom_cdf_discrete()`. The following example plots the $\text{Binomial}(10, 0.5)$ survival function.

```{r discrete-survival}
ggplot() +
  geom_survival_discrete(fun = dbinom, xlim = c(0, 10), args = list(size = 10, prob = 0.5))
```

**Hiding points.** Setting `show_points = FALSE` removes the endpoint circles.

```{r discrete-survival-no-points}
ggplot() +
  geom_survival_discrete(fun = dbinom, xlim = c(0, 10), args = list(size = 10, prob = 0.5),
    show_points = FALSE)
```

**Explicit support.** The survival function of the sample mean of 10 iid $\text{Bernoulli}(0.3)$ draws.

```{r discrete-survival-support}
f_mean <- function(x, prob) dbinom(round(x * 10), size = 10, prob = prob)
ggplot() +
  geom_survival_discrete(fun = f_mean, support = seq(0, 1, by = 0.1), args = list(prob = 0.3))
```


### Quantile function: `geom_qf()` and `geom_qf_discrete()`

The quantile function $Q(p) = \inf\{x : F(x) \ge p\}$ inverts the CDF.

**Continuous (`geom_qf()`).** Evaluates a quantile function over the unit interval $(0, 1)$ and draws it as a curve. The following example plots the standard normal quantile function.

```{r geom-qf}
ggplot() +
  geom_qf(fun = qnorm)
```

**Discrete (`geom_qf_discrete()`).** Takes a PMF and renders the quantile function as a left-continuous step function on the unit interval, with closed circles at the bottom of each jump (the value is achieved) and open circles at the top (the next value is not yet reached). The following example plots the $\text{Binomial}(10, 0.5)$ quantile function.

```{r discrete-qf}
ggplot() +
  geom_qf_discrete(fun = dbinom, xlim = c(0, 10), args = list(size = 10, prob = 0.5))
```

**Parameterized families.** Here we plot a $\text{Poisson}(5)$ discrete quantile function.

```{r discrete-qf-pois}
ggplot() +
  geom_qf_discrete(fun = dpois, xlim = c(0, 15), args = list(lambda = 5))
```

**Hiding points.** Setting `show_points = FALSE` removes the endpoint circles, leaving only the horizontal and vertical lines.

```{r discrete-qf-no-points}
ggplot() +
  geom_qf_discrete(fun = dbinom, xlim = c(0, 10), args = list(size = 10, prob = 0.5),
    show_points = FALSE)
```

**Hiding vertical lines.** Setting `show_vert = FALSE` removes the dashed vertical jump segments, leaving only the horizontal steps and endpoint circles.

```{r discrete-qf-no-vert}
ggplot() +
  geom_qf_discrete(fun = dbinom, xlim = c(0, 10), args = list(size = 10, prob = 0.5),
    show_vert = FALSE)
```

**Explicit support.** The quantile function of the sample mean of 10 iid $\text{Bernoulli}(0.3)$ draws.

```{r discrete-qf-support}
f_mean <- function(x, prob) dbinom(round(x * 10), size = 10, prob = prob)
ggplot() +
  geom_qf_discrete(fun = f_mean, support = seq(0, 1, by = 0.1), args = list(prob = 0.3))
```


### Hazard function: `geom_hf()`

`geom_hf()` plots the hazard function $h(x) = f(x) / S(x)$, the instantaneous rate of failure at time $x$ conditional on survival to that point. It requires both a PDF (`pdf_fun`) and a CDF (`cdf_fun`). Shared parameters go in `args`; use `pdf_args` and `cdf_args` for overrides when the two functions have different parameterizations. The exponential distribution's constant hazard confirms the memoryless property.

```{r hazard}
ggplot() +
  geom_hf(pdf_fun = dexp, cdf_fun = pexp, xlim = c(0.01, 10), args = list(rate = 0.5))
```

## Getting help

- Browse the [vignette](vignettes/ggfunction.Rmd) for a guided walkthrough
- File bugs or feature requests at <https://github.com/dusty-turner/ggfunction/issues>



## Installation

```{r installation, eval=FALSE}
# install.packages("pak")
pak::pak("dusty-turner/ggfunction")
```
