---
title: "Simulation Verification of KS Confidence Band Coverage"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Simulation Verification of KS Confidence Band Coverage}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment  = "#>",
  fig.width  = 7,
  fig.height = 4
)
library(ggfunction)
library(ggplot2)
```

`geom_ecdf()` and `geom_eqf()` draw simultaneous confidence bands using the
Dvoretzky--Kiefer--Wolfowitz (DKW) inequality with Massart's (1990) tight
constant.  The half-width is

$$
\varepsilon_n = \sqrt{\frac{\log(2/\alpha)}{2n}}, \qquad \alpha = 1 - \texttt{level},
$$

giving the band $[\hat{F}_n(x) - \varepsilon_n,\; \hat{F}_n(x) + \varepsilon_n]$
clipped to $[0, 1]$.  This band is guaranteed to contain the true CDF
everywhere simultaneously with probability **at least** $1 - \alpha$.

This vignette verifies the guarantee by simulation.

## What "simultaneous coverage" means

A band has **simultaneous** $(1-\alpha)$ coverage if

$$
P\!\left(\sup_{x} \lvert \hat{F}_n(x) - F(x) \rvert \leq \varepsilon_n\right)
\;\geq\; 1 - \alpha.
$$

The supremum equals the Kolmogorov–Smirnov statistic $D_n$, so we need
$P(D_n \leq \varepsilon_n) \geq 1 - \alpha$.  Because $F$ is continuous, $D_n$
has a distribution-free null distribution, so coverage does not depend on the
true distribution.

## Simulation

For each combination of sample size $n$ and nominal level $1 - \alpha$ we:

1. Draw $B = 10{,}000$ independent samples from $N(0, 1)$.
2. Compute $D_n = \sup_x \lvert \hat{F}_n(x) - F(x) \rvert$ via `ks.test()`.
3. Compute $\varepsilon_n$ using the same formula as `geom_ecdf()`.
4. Record the fraction of replications where $D_n \leq \varepsilon_n$.

```{r sim, cache = FALSE}
set.seed(20240101)

B      <- 10000
ns     <- c(10, 20, 50, 100, 200, 500, 1000)
levels <- c(0.90, 0.95, 0.99)

eps_fn <- function(n, level) sqrt(log(2 / (1 - level)) / (2 * n))

results <- do.call(rbind, lapply(ns, function(n) {
  do.call(rbind, lapply(levels, function(lv) {
    eps <- eps_fn(n, lv)
    dn  <- replicate(B, ks.test(rnorm(n), "pnorm", exact = FALSE)$statistic)
    data.frame(
      n         = n,
      level     = lv,
      nominal   = lv,
      empirical = mean(dn <= eps)
    )
  }))
}))
```

## Results table

Each cell shows empirical coverage (should be $\geq$ nominal).

```{r table, echo = FALSE}
wide <- reshape(
  results[, c("n", "level", "empirical")],
  idvar     = "n",
  timevar   = "level",
  direction = "wide"
)
colnames(wide) <- c("n", "90%", "95%", "99%")
wide[, 2:4] <- lapply(wide[, 2:4], function(x) paste0(round(x * 100, 1), "%"))
knitr::kable(wide, align = "rrrr",
             caption = paste0("Empirical simultaneous coverage over ", B,
                              " simulations from N(0,1)."))
```

Coverage is always **at or above** the nominal level. The bands are slightly
conservative for small $n$ (where the DKW bound is not yet tight) and
approach the nominal level as $n$ grows.

## Coverage plot

```{r plot, echo = FALSE}
results$label <- paste0(results$level * 100, "% nominal")

ggplot(results, aes(x = n, y = empirical, colour = factor(level),
                    group = factor(level))) +
  geom_hline(aes(yintercept = level, colour = factor(level)),
             linetype = "dashed", linewidth = 0.4) +
  geom_line(linewidth = 0.8) +
  geom_point(size = 2) +
  scale_x_log10(breaks = ns) +
  scale_y_continuous(labels = scales::percent_format(accuracy = 0.1),
                     limits = c(0.88, 1.0)) +
  scale_colour_manual(
    values = c("0.9" = "steelblue", "0.95" = "firebrick", "0.99" = "forestgreen"),
    labels = c("0.9" = "90%", "0.95" = "95%", "0.99" = "99%"),
    name   = "Nominal level"
  ) +
  labs(
    title    = "Empirical vs. nominal simultaneous coverage",
    subtitle = paste0("Dashed lines = nominal; points = empirical (B = ", B, " sims)"),
    x        = "Sample size n  (log scale)",
    y        = "Empirical coverage"
  ) +
  theme_minimal()
```

Empirical coverage (solid lines) lies above each dashed nominal line across
all $n$.  As $n$ increases the bands tighten toward the nominal level,
confirming that the DKW construction is asymptotically exact.

## Distribution-free check

Because $D_n$ under a continuous $F$ is distribution-free, the same coverage
holds for any continuous distribution.  A quick cross-check with Uniform and
Exponential confirms this:

```{r dist-free, cache = FALSE}
set.seed(20240102)

n   <- 100
lv  <- 0.95
eps <- eps_fn(n, lv)
B2  <- 10000

dists <- list(
  Normal      = function() rnorm(n),
  Uniform     = function() runif(n),
  Exponential = function() rexp(n),
  Beta        = function() rbeta(n, 2, 5)
)

dist_results <- do.call(rbind, lapply(names(dists), function(nm) {
  rfun   <- dists[[nm]]
  # Use the corresponding theoretical CDF
  pfun   <- switch(nm,
    Normal      = function(x) pnorm(x),
    Uniform     = function(x) punif(x),
    Exponential = function(x) pexp(x),
    Beta        = function(x) pbeta(x, 2, 5)
  )
  dn <- replicate(B2, {
    x  <- sort(rfun())
    fn <- seq_along(x) / n
    # Kolmogorov-Smirnov statistic: max over pre- and post-jump
    max(c(abs(fn - pfun(x)), abs((seq_along(x) - 1) / n - pfun(x))))
  })
  data.frame(distribution = nm, empirical = mean(dn <= eps))
}))

dist_results$nominal <- paste0(lv * 100, "%")
dist_results$empirical_pct <- paste0(round(dist_results$empirical * 100, 1), "%")
knitr::kable(dist_results[, c("distribution", "nominal", "empirical_pct")],
             col.names = c("Distribution", "Nominal", "Empirical"),
             align = "lcc",
             caption = paste0("Coverage at n = ", n, ", level = ", lv,
                              " across distributions (B = ", B2, " sims)."))
```

Coverage is at or above 95% for all distributions, confirming the
distribution-free guarantee.

## Conclusion

The simulations confirm that `geom_ecdf()` and `geom_eqf()` produce valid
simultaneous confidence bands: empirical coverage is always at or above the
nominal level, and the bands converge toward the nominal level as $n$ grows.
The DKW construction is sound for all tested sample sizes and distributions.
