---
title: "Empirical Geoms"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Empirical Geoms}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 6,
  fig.height = 4
)
```

```{r setup}
library(ggplot2)
library(ggfunction)
```

Empirical layers visualize samples and fitted distribution diagnostics without
requiring precomputed summaries.

```{r data}
set.seed(42)
x <- c(rnorm(80), rnorm(20, mean = 2))
df <- data.frame(x = x)
```

The empirical CDF can include DKW/Massart simultaneous bands for a fully
specified distribution-free ECDF band.

```{r ecdf}
ggplot(df, aes(x = x)) +
  geom_ecdf(conf_int = TRUE, level = 0.95)
```

The empirical quantile function uses the same sample but swaps the axes to show
sample quantiles against plotting positions.

```{r eqf}
ggplot(df, aes(x = x)) +
  geom_eqf(conf_int = TRUE, level = 0.95)
```

PP, QQ, and stabilized probability plots compare data with a fully specified
null distribution. Bands are DKW-based; when parameters are estimated from the
same data, interpret them as informal diagnostics.

```{r qq}
ggplot(df, aes(x = x)) +
  geom_qqplot(fun = qnorm, colour = "black") +
  coord_equal()
```

Censored survival summaries are available through Kaplan-Meier and
Nelson-Aalen-style empirical cumulative hazard layers.

```{r survival}
surv_df <- data.frame(
  time = c(1, 2, 2, 3, 4, 4, 5, 7),
  status = c(1, 1, 0, 1, 0, 1, 1, 0)
)

ggplot(surv_df, aes(x = time, status = status)) +
  geom_ecdf_km(conf_int = TRUE)

ggplot(surv_df, aes(x = time, status = status)) +
  geom_echf_na(conf_int = TRUE)
```

