---
title: "Numerical Conversions"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Numerical Conversions}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 6,
  fig.height = 4
)
```

```{r setup}
library(ggplot2)
library(ggfunction)
```

Theoretical distribution geoms accept several equivalent characterizations:
PDFs, CDFs, survival functions, quantile functions, hazards, and cumulative
hazards. When a requested layer is not supplied directly, `ggfunction` derives
it numerically using a shared set of conversion helpers.

The most important argument for numerical routes is `support`. It tells the
package where the distribution lives; `xlim` only controls the visible drawing
window.

```{r cdf-from-pdf}
ggplot() +
  geom_cdf(
    pdf_fun = dbeta,
    support = c(0, 1),
    xlim = c(0, 1),
    args = list(shape1 = 2, shape2 = 5)
  )
```

CDF-to-PDF routes use finite differences, with one-sided differences near
finite support boundaries.

```{r pdf-from-cdf}
ggplot() +
  geom_pdf(
    cdf_fun = pbeta,
    support = c(0, 1),
    xlim = c(0, 1),
    args = list(shape1 = 2, shape2 = 5)
  )
```

Hazard routes integrate from `hf_lower`; when `support` begins at a finite
lower endpoint, that endpoint is used as the default hazard origin.

```{r hazard}
h_weibull <- function(x, shape, scale) {
  ifelse(x >= 0, (shape / scale) * (x / scale)^(shape - 1), 0)
}

ggplot() +
  geom_pdf(
    hf_fun = h_weibull,
    support = c(0, Inf),
    xlim = c(0, 5),
    args = list(shape = 1.8, scale = 2)
  )
```

Numerical conversions are approximations. Accuracy depends on smoothness,
tail behavior, support bounds, and grid resolution. The manual script
`inst/benchmarks/accuracy-distribution-conversions.R` computes fixed-grid
errors for the conversion routes used in examples and manuscript claims.

