---
title: "Probability Geoms"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Probability Geoms}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 6,
  fig.height = 4
)
```

```{r setup}
library(ggplot2)
library(ggfunction)
```

`ggfunction` separates the visible plotting window from the computational
support of a probability distribution. The `xlim` argument controls what is
drawn, while `support` controls numerical integration, inversion, validation,
and probability shading.

```{r normal-tail}
ggplot() +
  geom_pdf(fun = dnorm, xlim = c(-1, 1), p = 0.975) +
  labs(title = "The visible window clips a distributional tail probability")
```

For finite-support distributions, supply the support explicitly.

```{r beta-support}
ggplot() +
  geom_pdf(
    fun = dbeta,
    xlim = c(0, 1),
    support = c(0, 1),
    args = list(shape1 = 2, shape2 = 5),
    p_lower = 0.1,
    p_upper = 0.9
  )
```

Highest-density regions are grid approximations. Use `hdr_xlim` when the HDR
threshold should be computed on a wider interval than the visible window.

```{r hdr}
f_mix <- function(x) 0.55 * dnorm(x, -1.8, 0.45) + 0.45 * dnorm(x, 1.6, 0.35)

ggplot() +
  geom_pdf(
    fun = f_mix,
    xlim = c(-3, 3),
    hdr_xlim = c(-5, 5),
    shade_hdr = 0.8
  )
```

Discrete distributions use exact support points. For HDRs, tied masses at the
cutoff are included, so achieved coverage can exceed the requested value.

```{r discrete}
ggplot() +
  geom_pmf(
    fun = dbinom,
    support = 0:12,
    args = list(size = 12, prob = 0.35),
    shade_hdr = c(0.5, 0.8, 0.95)
  )
```

