---
title: "The sample Aesthetic in Goodness-of-Fit Layers"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{The sample Aesthetic in Goodness-of-Fit Layers}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 6,
  fig.height = 4
)
```

```{r setup}
library(ggplot2)
library(ggfunction)
```

ggfunction's one-sample goodness-of-fit layers --- `geom_ppplot()`,
`geom_qqplot()`, and `geom_spplot()` --- take their observations through the
**`sample`** aesthetic:

```{r first-example}
set.seed(1)
df <- data.frame(hours = rlnorm(60, meanlog = 1, sdlog = 0.6))

ggplot(df, aes(sample = hours)) +
  geom_qqplot(fun = qlnorm, args = list(meanlog = 1, sdlog = 0.6))
```

If you have used ggplot2's own `geom_qq()`, this is the same convention:
`ggplot(df, aes(sample = hours)) + geom_qq()`. But if you are used to writing
`aes(x = hours)` for histograms, density plots, and ECDFs, `sample` can look
arbitrary. It is not. This vignette explains, as concretely as possible, the
failure mode that `sample` exists to prevent.

The short version: **`x` and `y` are position aesthetics, and ggplot2
transforms position aesthetics *before* a layer's statistic runs.** A
goodness-of-fit statistic must evaluate the null distribution at the *raw*
observations, and those observations are not themselves a plotted position ---
in a PP plot, neither axis displays the data at all. Mapping the input to a
position aesthetic therefore hands the statistic corrupted values whenever a
position scale carries a transformation. `sample` is a non-position aesthetic,
so ggplot2 never touches it, and the statistic is guaranteed to see your data
exactly as you supplied it.

The rest of this vignette walks through that argument step by step.

## What ggplot2 does to aesthetics before a stat runs

Every ggplot2 layer computes in a pipeline: aesthetics are evaluated against
the data, then *position* aesthetics (`x`, `y`, and friends) are passed
through their scales' transformations, and only then does the layer's
statistic see the data. Non-position aesthetics --- `colour`, `weight`,
`group`, and ggfunction's `sample` --- skip the transformation step entirely.

You can see the pre-stat transformation directly with a stat that simply
reports what it received:

```{r pipeline-demo}
StatShow <- ggproto("StatShow", Stat,
  required_aes = "x",
  compute_group = function(data, scales) {
    message("the stat received x = ", paste(round(data$x, 3), collapse = ", "))
    data
  }
)

d <- data.frame(x = c(1, 10, 100), y = 1:3)

# identity scale: the stat sees the data as-is
invisible(ggplot_build(
  ggplot(d, aes(x, y)) + layer(
    stat = StatShow, geom = "point", position = "identity", params = list()
  )
))

# log10 scale: the stat sees log10(data), *not* the data
invisible(ggplot_build(
  ggplot(d, aes(x, y)) + layer(
    stat = StatShow, geom = "point", position = "identity", params = list()
  ) + scale_x_log10()
))
```

For most statistics this is exactly right. A boxplot of log-scaled data
*should* compute medians and hinges in log space; that is what "putting the
axis on a log scale" means for descriptive summaries.

## Why that behavior breaks goodness-of-fit diagnostics

A one-sample diagnostic is different from a descriptive summary in one
crucial way: it evaluates a *user-specified null distribution* at the
observations. The PP plot computes $F_0(x_{(i)})$ against the plotting
positions $p_i$; the QQ plot computes $Q_0(p_i)$ against the order statistics
$x_{(i)}$; the SP plot variance-stabilizes the PP coordinates. In every case
the null function `fun` must receive the observations **in the units the null
distribution is defined on** --- the raw data.

Now suppose the observations were mapped to `x`, and consider the most
natural thing a user of the lognormal data above might do: add
`scale_x_log10()` and `scale_y_log10()` to a QQ plot, so that the heavily
skewed points spread out legibly. That is a pure *display* request --- the
user wants the same diagnostic drawn on log axes. But because `x` is a
position aesthetic, ggplot2 would transform the observations *before* the
statistic ran, and the null quantile function would be compared against
$\log_{10}(x_{(i)})$ instead of $x_{(i)}$.

We can compute by hand exactly what that mis-specified comparison looks like,
next to the correct one. The data below are drawn from the null itself, so a
truthful diagnostic must hug the identity line:

```{r corruption-demo, fig.width = 7, fig.height = 3.5}
x_sorted <- sort(df$hours)
p <- ppoints(length(x_sorted), a = 1 / 2)
theoretical <- qlnorm(p, meanlog = 1, sdlog = 0.6)

correct <- data.frame(
  theoretical = theoretical,
  observed = x_sorted,          # raw observations: what the stat must see
  computation = "correct: fun receives raw data"
)
corrupted <- data.frame(
  theoretical = theoretical,
  observed = log10(x_sorted),   # what a pre-stat log10 x scale would deliver
  computation = "corrupted: fun receives log10(data)"
)

ggplot(rbind(correct, corrupted), aes(theoretical, observed)) +
  geom_abline(slope = 1, intercept = 0, linetype = "dashed") +
  geom_point() +
  facet_wrap(~computation, scales = "free_y") +
  labs(x = "theoretical quantiles", y = "observed values")
```

The left panel is the diagnostic; the right panel is what the same layer
would silently draw if its input rode through a transformed position scale.
The data were simulated *from the null*, yet the right panel "rejects" it
decisively. Nothing about the plot warns you that this happened --- the axes
are labeled, the points are crisp, and the conclusion is garbage. Silent
corruption of a *diagnostic* is about the worst failure a plotting layer can
have, because the entire purpose of the layer is to be evidence.

There is no way to repair this inside the statistic. Once the positions have
been transformed, the stat cannot reliably recover the raw values (it would
need to know which transformation ran, invert it exactly, and trust that no
information was lost --- none of which ggplot2's stat API promises).

### A second, quieter problem: one aesthetic, two meanings

Even on identity scales, `aes(x = observations)` overloads the aesthetic.
Consider what the x axis *displays* in each plot:

| layer           | input to `fun`  | x axis displays            | y axis displays          |
|-----------------|-----------------|----------------------------|--------------------------|
| `geom_ppplot()` | raw observations| theoretical probabilities  | observed probabilities   |
| `geom_spplot()` | raw observations| stabilized probabilities   | stabilized probabilities |
| `geom_qqplot()` | raw observations| theoretical quantiles      | observed values          |

In a PP plot the observations appear on *neither* axis. With the observations
mapped to `x`, the single x scale would simultaneously control two unrelated
things: the units of the data entering the computation (before the stat) and
the styling of the theoretical-probability axis (after it). A user who writes
`scale_x_continuous(labels = scales::percent)` means the probability axis; a
user who writes `scale_x_log10()` on lognormal data means the input. The
grammar gives them one knob for both. Separating the input (`sample`) from
the display positions (whatever the stat computes) dissolves the ambiguity.

## What `sample` guarantees

Because `sample` is not a position aesthetic, it is never transformed by any
scale. The layers therefore have a clean, checkable contract:

1. the null CDF or quantile function **always receives raw observations and
   raw probabilities**, and
2. the computed display coordinates are transformed **exactly once** by the
   output scales.

So transformed axes now do precisely what the user means --- they restyle the
display without touching the computation. The log-log QQ plot that motivated
this vignette:

```{r qq-log-log}
ggplot(df, aes(sample = hours)) +
  geom_qqplot(fun = qlnorm, args = list(meanlog = 1, sdlog = 0.6)) +
  scale_x_log10() +
  scale_y_log10()
```

You can verify the contract on a sample chosen to make the arithmetic
readable. With observations 1, 10, 100 under a standard lognormal null, the
raw values survive into the computed data (`after_stat(sample)`), the
plotting positions are the exact ppoints, and only the *positions* carry the
log transformation:

```{r contract-check}
tiny <- data.frame(value = c(1, 10, 100))
b <- ggplot_build(
  ggplot(tiny, aes(sample = value)) +
    geom_qqplot(fun = qlnorm, a = 0.5, conf_int = FALSE, identity_line = FALSE) +
    scale_x_log10() +
    scale_y_log10()
)$data[[1]]

b$sample      # raw observations, untouched: 1, 10, 100
b$p           # raw plotting positions: 1/6, 1/2, 5/6
b$y           # display positions: log10 of the observations = 0, 1, 2
all.equal(b$x, qnorm(ppoints(3, a = 0.5)) / log(10))  # log10(qlnorm(p))
```

This is also why the confidence bands can be trusted under transformed
scales: the band is computed from raw probabilities and raw quantiles, and
its ribbon coordinates are transformed once for display, the same as the
points.

## The legacy `aes(x = )` input

Earlier versions of these layers took `aes(x = observations)`, and plenty of
existing code does so. That spelling still works wherever it is safe --- an
identity x scale --- with a deprecation warning pointing here:

```{r legacy-warning, warning = TRUE}
p_legacy <- ggplot(df, aes(x = hours)) +
  geom_ppplot(fun = plnorm, args = list(meanlog = 1, sdlog = 0.6),
              conf_int = FALSE)
invisible(ggplot_build(p_legacy))
```

Under a transformed x scale there is no safe interpretation, for the reasons
demonstrated above, so the layer refuses rather than draw the corrupted
diagnostic:

```{r legacy-error, error = TRUE}
p_bad <- ggplot(df, aes(x = hours)) +
  geom_ppplot(fun = plnorm, args = list(meanlog = 1, sdlog = 0.6),
              conf_int = FALSE) +
  scale_x_log10()
ggplot_build(p_bad)
```

An error where the old behavior silently drew a wrong plot is the point: the
one thing a goodness-of-fit layer must never do is lie.

## Summary

- `geom_ppplot()`, `geom_qqplot()`, and `geom_spplot()` take observations via
  `aes(sample = )`, following `ggplot2::geom_qq()`.
- Position aesthetics are transformed before statistics run; a null
  distribution evaluated at transformed observations produces a silently
  wrong diagnostic. `sample` is immune because it is not a position.
- Output scales (`scale_x_log10()`, `scale_y_reverse()`, ...) restyle the
  display only; computations happen on raw values, and raw computed columns
  (`after_stat(sample)`, `after_stat(p)`, `after_stat(theoretical)`,
  `after_stat(observed)`) remain available.
- `aes(x = )` remains supported on identity scales for backward
  compatibility, with a deprecation warning; under a transformed x scale it
  errors instead of corrupting the input.
