---
title: "ggfunction: Functions and Distributions in ggplot2"
abstract: >
  The ggfunction package provides ggplot2 layers for plotting
  mathematical functions, probability distributions, empirical summaries, and
  censored-data summaries directly from function definitions or samples. Its
  design has three parts. First, a dimensional taxonomy covers scalar
  functions, parametric curves, scalar fields, and vector fields. Second,
  probability layers display density, mass, distribution, survival, quantile,
  hazard, and cumulative-hazard functions with support-aware shading and
  documented conversion routes among representations. Third, empirical layers
  compute ECDFs, empirical quantiles, PP/QQ/SP diagnostics, empirical mass
  functions, empirical cumulative hazards, Kaplan--Meier survival curves, and
  Nelson--Aalen cumulative hazards. The displays are standard ggplot2 stats and
  geoms, so they compose with scales, coordinates, themes, facets, and
  annotations. Confidence displays are documented per layer, and fitted-null
  diagnostic bands are exploratory rather than calibrated unless separately
  adjusted.
author:
  - name: Dusty Turner
    affiliation: United States Military Academy
    address: West Point, NY
    email: dusty.s.turner@gmail.com
  - name: David Kahle
    affiliation: Baylor University
    address: Waco, TX
    email: david@kahle.io
    orcid: 0000-0002-9999-1558
  - name: Rodney X. Sturdivant
    affiliation: Baylor University
    address: Waco, TX
    email: rodney_sturdivant@baylor.edu
  - name: James Otto
    affiliation: Alcon Inc.
    address: Fort Worth, TX
    email: jamesotto852@gmail.com
    note: Early drafts of the (E)CDF stats and geoms.
date: "2026-06-28"
date_received: ~
journal:
  firstpage: 1
  lastpage: ~
slug: turner-kahle-sturdivant-ggfunction
creative_commons: CC BY
packages:
  cran:
    - ggfunction
    - ggplot2
    - ggvfields
    - ggdist
    - ggdensity
    - qqplotr
    - survminer
    - ggsurvfit
    - ggformula
    - hdrcde
    - mosaic
    - mosaicCalc
    - metR
    - rlang
    - cli
    - patchwork
    - bench
    - roxygen2
    - lattice
  bioc: ~
  other: ~
CTV: ~
output:
  rjtools::rjournal_pdf_article:
    toc: no
  rjtools::rjournal_web_article:
    self_contained: yes
    toc: no
    includes:
      in_header: resources/mathjax_macros.html
    lua-filter: parse-fig-caption.lua
    dev: svg
bibliography: RJreferences.bib
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = FALSE,
  collapse = TRUE,
  comment = "#>",
  fig.width = 6,
  fig.height = 3,
  fig.align = "center",
  out.width = "85%",
  warning = FALSE,
  message = FALSE
)
```

# Introduction

Many statistical graphics begin with a function rather than a data frame. An
instructor shades the upper tail of a $t$ distribution, a student compares an
empirical CDF with a fitted model, or an analyst checks whether a censored
sample is consistent with an exponential survival curve. In base \proglang{R}
[@R],
these tasks are straightforward but often require hand-built grids, numerical
integration, and ad hoc annotations. In \CRANpkg{ggplot2}
[@wickham2010layered; @wickham2016ggplot2], the built-in
`stat_function()` covers only the univariate case $f:\mathbb{R}\to
\mathbb{R}$ and leaves probability-specific calculations to the user.

\pkg{ggfunction} addresses this gap by making functions layer-level objects in
\CRANpkg{ggplot2}. Users supply an \proglang{R} function, optional parameters in
`args`, and a domain. The layer evaluates the function, performs any
requested statistical transformation, and returns ordinary \CRANpkg{ggplot2}
data for rendering. Figure \@ref(fig:motivating) illustrates the motivating
case: the user asks for a distributional tail probability, and the layer
handles the quantile lookup, support-aware shading, and \CRANpkg{ggplot2}
display.

Two design choices distinguish \pkg{ggfunction} from small plotting wrappers.
The probability layers share documented conversion routes among density or
mass, distribution, survival, quantile, hazard, and cumulative-hazard
representations, so one supplied characterization can drive several views of
the same distribution. The package also separates statistical support from the
display window: support governs normalization, integration, inversion, and
probability shading, while `xlim` and coordinates govern what is visible.
Together, theoretical curves, empirical summaries, censored-data estimators,
and diagnostic plots use the same function-as-layer pattern while remaining
ordinary \CRANpkg{ggplot2} layers. The result is a practical, composable
interface for common function- and distribution-based graphics in \CRANpkg{ggplot2}.

The package is available from CRAN and can be installed with
`install.packages("ggfunction")`. Development versions and issue tracking are
available from \url{https://github.com/dusty-turner/ggfunction}. The package
depends on \CRANpkg{ggplot2}; bivariate highest-density regions are delegated
to \CRANpkg{ggdensity} [@otto2023ggdensity], and vector-field and stream-field
layers delegate to \CRANpkg{ggvfields} [@ggvfields].

The rest of the article is organized around package design and workflows rather than a complete reference manual; full function-level documentation is available on the package website and in the vignettes. Section 2 describes the layer design and
dimensional taxonomy. Section 3 presents three core workflows: mathematical
functions, probability distributions, and empirical or censored data summaries.
Section 4 summarizes implementation, numerical behavior, validation, and
performance. Sections 5--7 discuss composability, related work, limitations, and
future directions.

# Design

## Functions as layer-level objects

\pkg{ggfunction} does not change the grammar of graphics itself. It supplies new
`ggplot2::layer()` constructors. Each constructor combines a `Stat`
that evaluates a function or computes an empirical summary with a `Geom`
that renders the returned data. This follows the \CRANpkg{ggplot2} extension
model: the result is a standard layer that can be added with `+`, trained
by ordinary scales, clipped by coordinates, faceted, themed, and annotated.

The key interface choice is that functions are layer parameters, not aesthetics.
This is deliberate. A function such as `dnorm()` or a bivariate density is
not a column in the plot data nor is it a visual property of a geometric object; it is the object the stat evaluates over other aesthetics (typically the spatial ones `x` and `y`) to create
plot data. Additional function parameters are supplied through `args`,
mirroring `ggplot2::stat_function()` and avoiding anonymous wrappers. The basic specification syntax looks like this; the resulting plot is in Figure \@ref(fig:motivating):

```{r motivating, fig.height=3, fig.cap="A motivating example: support-aware normal-density plotting and shading with a simple syntax. The shaded region is computed from the distribution, not by renormalizing over the displayed x range.", echo=TRUE, message=FALSE}
library("ggplot2"); theme_set(theme_minimal())
theme_update(panel.grid.minor = element_blank())
library("ggfunction")

ggplot() +
  geom_pdf(
    fun = dnorm, xlim = c(-3.5, 3.5),
    support = c(-Inf, Inf), p_lower = 0.025, p_upper = 0.975
  )
```

This design keeps the common case short while preserving explicit control over
the domain, grid resolution, and rendering choices.

## Dimensional taxonomy

The mathematical-function family is organized by domain and codomain dimension.
Table \@ref(tab:taxonomy-table) gives the taxonomy, and Figure
\@ref(fig:taxonomy) shows the corresponding displays.
The names encode the function signature. That convention is more verbose than
`geom_function()`, but it makes the dimensional assumptions visible at the
call site. Probability and empirical layers use statistical names instead:
`geom_pdf()`, `geom_ecdf()`, `geom_ecdf_km()`, and so on.

```{r taxonomy-table}
#| tab.cap: "Dimensional taxonomy for mathematical-function layers."
taxonomy <- data.frame(
  Signature = c(
    "$f:\\mathbb{R}\\to\\mathbb{R}$",
    "$\\boldsymbol{\\gamma}:\\mathbb{R}\\to\\mathbb{R}^2$",
    "$f:\\mathbb{R}^2\\to\\mathbb{R}$",
    "$\\mathbf{F}:\\mathbb{R}^2\\to\\mathbb{R}^2$"
  ),
  Layer = c(
    "\\code{geom\\_function\\_1d\\_1d()}",
    "\\code{geom\\_function\\_1d\\_2d()}",
    "\\code{geom\\_function\\_2d\\_1d()}",
    "\\code{geom\\_function\\_2d\\_2d()}"
  ),
  Display = c(
    "curve, optionally shaded",
    "parametric path with parameter colour",
    "raster, contours, or filled contours",
    "arrows or streamlines"
  ),
  check.names = FALSE
)
knitr::kable(taxonomy, escape = FALSE)
```

```{r taxonomy, fig.width=7, fig.height=5.8, out.width="95%", fig.cap="The four mathematical-function signatures supported by the ggfunction dimensional taxonomy: scalar functions, parametric curves, scalar fields, and vector fields."}
scalar_plot <- ggplot() +
  geom_function_1d_1d(fun = sin, xlim = c(0, 2 * pi)) +
  labs(title = expression(R %->% R), x = "x", y = "sin(x)")

circle <- function(t) c(cos(t), sin(t))
curve_plot <- ggplot() +
  geom_function_1d_2d(fun = circle, tlim = c(0, 2 * pi), tail_point = TRUE) +
  coord_equal() +
  labs(title = expression(R %->% R^2), x = "x(t)", y = "y(t)")

field <- function(v) sin(v[1]) * cos(v[2])
scalar_field_plot <- ggplot() +
  geom_function_2d_1d(fun = field, xlim = c(-pi, pi), ylim = c(-pi, pi), n = 60, type = "contour_filled") +
  labs(title = expression(R^2 %->% R), x = "x", y = "y", fill = "f")

rotation <- function(v) c(-v[2], v[1])
vector_plot <- ggplot() +
  geom_function_2d_2d(fun = rotation, xlim = c(-1, 1), ylim = c(-1, 1), n = 9, type = "vector") +
  coord_equal() +
  labs(title = expression(R^2 %->% R^2), x = "x", y = "y")

library("patchwork")
(scalar_plot + curve_plot) / (scalar_field_plot + vector_plot)
```



# Three workflows

## Mathematical functions by dimension

For mathematical functions, the user's main decisions are the signature, the
domain, and the visual encoding. A scalar function is evaluated on an equally
spaced sequence over `xlim`. A parametric curve is evaluated over `tlim`; the
returned length-two vector supplies the path coordinates. A
scalar field is evaluated on an $n_x \times n_y$ grid and then handed to the
corresponding \CRANpkg{ggplot2} raster or contour machinery. A vector field is
evaluated at seed points and then drawn by \CRANpkg{ggvfields}; stream fields also
delegate integration settings such as `method`, `max_it`, `T`,
and `L`.

This workflow is useful when the data of interest are generated by a known
mathematical rule. It also gives probability and diagnostic layers a common
implementation pattern: create the grid, evaluate a function or summary, and
return ordinary plot data.

The taxonomy is intentionally small. It covers the function signatures that can
be displayed naturally in a two-dimensional plotting window without a separate
three-dimensional graphics system. A scalar field, for example, can be
displayed by mapping its output to colour or contour level while retaining the
two input coordinates as the horizontal and vertical axes. A vector field can
be displayed by drawing short vectors or streamlines at selected seed points.
Higher-dimensional functions require additional choices--conditioning,
projection, slicing, animation, or interactivity--that fall outside this
package's current scope.

Two implementation details matter in this workflow. First, evaluation is
grid-based. Increasing `n` improves smoothness and contour resolution, but it
also increases computation, especially for two-dimensional grids. Second,
`args` is spliced into the user's function at evaluation time. This lets a
single function represent a parametric family without redefining closures, for example:

```{r dimension-args-example, eval=FALSE, echo=TRUE}
lissajous <- function(t, a = 3, b = 2, delta = pi / 2) c(sin(a*t + delta), sin(b*t))

ggplot() +
  geom_function_1d_2d(fun = lissajous, tlim = c(0, 2 * pi), args = list(a = 5, b = 4))
```

For scalar and vector fields, the same convention means that model parameters,
physical constants, or tuning values can remain explicit in the plotting call.
This is especially helpful in teaching settings, where the relationship between
parameters and shape is often the point of the graphic. It also helps avoid
subtle reproducibility problems: the plot source records both the mathematical
function and the parameter values used to evaluate it.

Vector fields can be rendered either as local arrows or as streamlines. Both
views start from the same vector-valued function, but they answer different
visual questions. Arrows emphasize local direction and magnitude at selected
grid points; streamlines emphasize integral curves of the field. In
\pkg{ggfunction}, both displays are delegated to \CRANpkg{ggvfields}.


## Probability distributions

The probability workflow starts from one characterization of a distribution and
asks for another visual representation. For example, users may have a density
but want a CDF, a CDF but want a quantile function, or a hazard function but
want a density. Native inputs are preferred when available as `fun`. When a
native input is absent, \pkg{ggfunction} can derive the needed function through
the routes shown in Tables \@ref(tab:conversion-continuous) and
\@ref(tab:conversion-discrete); the input argument name signals the function
type, e.g. supplying `hf_fun` to `geom_pdf()`.

The continuous one-dimensional layers use the standard identities among
distribution functions. If $f$ is a density and $F$ is the CDF, then
\[
  F(x) = \int_{-\infty}^{x} f(t)\,dt,\qquad
  S(x) = 1-F(x),\qquad
  Q(p) = F^{-1}(p).
\]
For survival and reliability displays, the hazard and cumulative hazard are
\[
  h(x) = \frac{f(x)}{S(x)},\qquad
  H(x) = -\log S(x),\qquad
  S(x) = \exp\{-H(x)\}.
\]
The package does not rederive these relationships for the reader each time a
geom is used. Instead, it encodes them in conversion helpers that are shared by
the distribution layers. This keeps the user-facing API small while making the
route explicit in documentation and validation tests.

Discrete distributions use analogous operations on support points. A PMF is
accumulated by summation rather than integration, and a quantile function is a
generalized inverse over the ordered support. Because discrete supports can be
finite, infinite, integer-valued, or explicitly supplied, the `support` argument
is more than a plotting convenience. Supplied support values are sorted before
PMF values are accumulated; the vector is not treated as a traversal order. For
example, if the mass at 2 is \(1/2\), the mass at 1 is \(1/3\), and the mass at
3 is \(1/6\), then `support = c(2, 1, 3)` still yields CDF jumps at 1, then 2,
then 3. The CDF jumps from 0 to \(1/3\) at 1, to \(1/3+1/2\) at 2, and to 1 at
3. This sorted support determines where step functions jump and how probability
shading is resolved. For discrete cumulative targets that cannot be hit exactly,
the discrete layers---`geom_pmf()` and the step geoms `geom_cdf_discrete()`,
`geom_survival_discrete()`, and `geom_qf_discrete()`---include the first
support point whose cumulative mass reaches or exceeds the target; discrete
HDR shading similarly uses the smallest containing set, including all support
points tied at the cutoff mass.

The bivariate probability layers have a narrower scope. `geom_pdf_2d()` draws
theoretical bivariate densities as HDRs or rasters, and `geom_pmf_2d()` draws
bivariate mass functions on a lattice. The package does not estimate bivariate
densities from samples; that task belongs to \CRANpkg{ggdensity}, which follows the same general design philosophy. It also does
not provide arbitrary three-dimensional rendering of density surfaces. The goal
is to display probability content in the same two-dimensional grammar used by
the rest of \CRANpkg{ggplot2}.

```{r conversion-continuous}
#| tab.cap: "Continuous distribution-function inputs and conversion routes. Native inputs are used directly; derived routes may require numerical integration, finite differences, interpolation, or root finding."
conversion_continuous <- data.frame(
  Target = c(
    "PDF $f$",
    "CDF $F$",
    "survival $S$",
    "quantile $Q$",
    "hazard $h$",
    "cum. hazard $H$"
  ),
  "Other sources" = c(
    "$F$, $S$, $Q$, or $h$",
    "$f$, $S$, $Q$, or $h$",
    "$F$, $f$, $Q$, or $h$",
    "$F$, $f$, $S$, or $h$",
    "$f$, $F$, $S$, or $Q$",
    "$h$, $F$, $S$, $f$, or $Q$"
  ),
  Route = c(
    "finite differences, interpolation, or $hS$",
    "\\code{stats::integrate()} or identity",
    "$1-F$",
    "bracketed \\code{stats::uniroot()} on $F$",
    "$f/S$ with tail-stable $S$; $-(\\log S)'$ from $S$; $f+F$ accepted",
    "$-\\log S$ or \\code{stats::integrate()} on $h$"
  ),
  check.names = FALSE
)
knitr::kable(conversion_continuous, escape = FALSE)
```

```{r conversion-discrete}
#| tab.cap: "Discrete distribution-function inputs and conversion routes. Computations are performed over the declared support."
conversion_discrete <- data.frame(
  Target = c("PMF", "CDF", "survival", "quantile"),
  "Other sources" = c(
    "CDF or survival",
    "PMF or survival",
    "PMF or CDF",
    "PMF, CDF, or survival"
  ),
  Route = c(
    "support differences",
    "cumulative sums over sorted support",
    "$1-F$ on sorted support",
    "generalized inverse over sorted support"
  ),
  check.names = FALSE
)
knitr::kable(conversion_discrete, escape = FALSE)
```

A single source can therefore drive several views of the same distribution.
Figure \@ref(fig:gamma-views) starts from the gamma density and asks
\pkg{ggfunction} to draw related functions. Two of the panels use native
functions, while the CDF and survival panels demonstrate PDF-derived routes.
In a manuscript or analysis script, this makes the distributional relationship
explicit in the plotting code. The design makes the code straightforward:

```{r gamma-views, fig.width=7, fig.height=5.8, out.width="95%", fig.cap="One gamma distribution viewed through four related functions. The PDF panel uses a native density; the CDF and survival panels derive from the density; the hazard panel uses the documented PDF + CDF bundle.", echo=TRUE}
gamma_args <- list(shape = 2, scale = 1.5)
Sx <- c(0, Inf)
gx <- c(0, 10)
hx <- c(0.01, 10)

gamma_pdf <- ggplot() +
  geom_pdf(fun = dgamma, args = gamma_args, xlim = gx, support = Sx) +
  labs(title = "PDF", x = "x", y = "f(x)")

gamma_cdf <- ggplot() +
  geom_cdf(pdf_fun = dgamma, args = gamma_args, xlim = gx, support = Sx) +
  labs(title = "CDF from PDF", x = "x", y = "F(x)")

gamma_surv <- ggplot() +
  geom_survival(pdf_fun = dgamma, args = gamma_args, xlim = gx, support = Sx) +
  labs(title = "Survival from PDF", x = "x", y = "S(x)")

gamma_hazard <- ggplot() +
  geom_hf(pdf_fun = dgamma, cdf_fun = pgamma,
          args = gamma_args, xlim = hx, support = Sx) +
  labs(title = "Hazard from PDF + CDF", x = "x", y = "h(x)")

library("patchwork")
(gamma_pdf + gamma_cdf) / (gamma_surv + gamma_hazard)
```

The distinction between support and display window is important. The `xlim`
argument controls what is drawn. The `support` argument controls distributional
calculations: normalization checks, PDF-to-CDF integration, CDF-to-quantile
inversion, and probability shading. For continuous one-dimensional probability
layers, omitting `support` leaves the computational support at its default
`c(-Inf, Inf)`; an `xlim` value alone does not redefine the distribution. Thus
a 97.5% normal upper-tail request is interpreted with respect to the full normal
distribution even if the visible window is narrower. Figure
\@ref(fig:prob-shading) shows lower-tail, central-interval, and two-tail
requests.

This separation is easy to miss because ordinary plotting code often treats
the x-axis limits as the whole domain. For probability displays, that would
change the meaning of shaded area. Suppose a density has support
$(-\infty,\infty)$ and the user displays only $[-2,2]$. If a lower-tail request
with $p=0.95$ were normalized over the visible window, the shaded boundary
would mark 95% of the mass inside $[-2,2]$, not the 95th percentile of the
distribution. That is rarely what a statistical diagram intends. In
\pkg{ggfunction}, quantiles, tail probabilities, and central intervals are
computed on the declared support; the coordinate system then determines which
part of the result is visible.

More generally, each distribution layer operates on the pair (function,
support), and the package maintains a one-way containment among three
domains: the declared support determines every statistical quantity; the
display window (`xlim`) selects which part of the result is visible; and the
scales' coordinate transformations style that view. Nothing flows backward.
All limits and supports are interpreted in data coordinates, and under
transformed scales the layers build their evaluation grids evenly in panel
space, evaluate the user's function at the data-space preimages, and
transform computed positions exactly once for display---a deliberate
difference from `ggplot2::stat_function()`, which interprets `xlim` in
transformed scale space. Each stat retains raw statistical values
(evaluation points, densities, cumulative probabilities, shading boundaries)
alongside display positions, so probability calculations never consume a
transformed number: a shading boundary is an exact distributional quantile,
invariant to the grid resolution and to the axis scales.

The same distinction matters for bounded supports. A beta density should be
integrated over $[0,1]$, not over the displayed portion of that interval. A
Weibull or exponential hazard should accumulate from time zero unless the user
specifies another origin. A custom distribution with support on a nonstandard
finite interval should declare that interval so numerical integration and
root-finding do not waste effort outside the distribution's domain or return
misleading boundary behavior. The `support` argument is therefore a statistical
parameter of the layer, not a cosmetic axis setting.

Discrete layers have a different fallback because their computation requires a
finite set of support points. If `support` is omitted and `xlim` is supplied,
the package evaluates the integer support from `ceiling(xlim[1])` through
`floor(xlim[2])`; if both are omitted, it uses a small default support. This is
convenient for quick binomial and Poisson examples, but it also means that
`xlim` is part of the discrete computation unless `support` is supplied. When
the drafted support demonstrably fails to carry the distribution's mass, the
layer alerts and proceeds with the declared, truncated object; in analyses
where `xlim` is only a visual crop, users should declare the full `support`
explicitly and crop with coordinates.

Display windows still have an important role. They let the user focus on the
region where a comparison is readable, crop extreme tails, or align several
plots with a common x-axis. The recommended pattern is to compute on the
distributional support and use the plotting window only for visual emphasis.
For HDR displays, the analogous pattern is to compute over `hdr_xlim` and
`hdr_ylim` and use `coord_cartesian()` if a narrower view is desired. This is
the same principle in one or two dimensions: probability content belongs to the
computation domain, while readability belongs to the display domain.

```{r prob-shading, fig.width=7, fig.height=3, out.width="95%", fig.cap="Three probability-shading requests for the standard normal distribution: a lower-tail probability, a central interval, and two tails outside a central interval."}
base_pdf <- function(...) {
  ggplot() +
    geom_pdf(fun = dnorm, xlim = c(-3.5, 3.5), ...) +
    labs(x = "x", y = "density")
}

p_lower <- base_pdf(p = 0.9) + ggtitle(expression(F(x) <= "0.90"))
p_central <- base_pdf(p_lower = 0.025, p_upper = 0.975) + ggtitle("central 95%")
p_outside <- base_pdf(p_lower = 0.025, p_upper = 0.975, shade_outside = TRUE) +
  ggtitle("two tails")

p_lower + p_central + p_outside
```

Highest density regions (HDRs) are threshold-based regions of largest density
containing a requested probability. In one dimension, \pkg{ggfunction} computes
an approximate threshold on a grid over `hdr_xlim`, using `xlim` only as the
fallback HDR computation interval when `hdr_xlim` is omitted. In two dimensions,
`geom_pdf_2d()` adapts the package's vector-argument density interface to
\CRANpkg{ggdensity}'s `geom_hdr_fun()` and `geom_hdr_lines_fun()` interfaces.
\CRANpkg{ggdensity} uses its own `xlim`/`ylim` arguments as the functional HDR
domain; \pkg{ggfunction} exposes `hdr_xlim` and `hdr_ylim` so users can choose a
larger HDR computation domain while keeping the displayed `xlim`/`ylim` narrow.
These HDRs are gridded approximations whose accuracy depends on the
computational domain and grid resolution. For a narrower view after computing
on a wider domain, users should compute over the larger `hdr_xlim`/`hdr_ylim`
and use `coord_cartesian()` for display.

```{r hdr-examples, fig.width=7, fig.height=3, out.width="95%", fig.cap="Approximate highest density regions. Left: a disconnected 80% grid-based HDR for a bimodal univariate density. Right: bivariate normal HDRs delegated to ggdensity."}
bimodal <- function(x) {
  0.6 * dnorm(x, mean = -2, sd = 0.6) +
    0.4 * dnorm(x, mean = 2, sd = 1.2)
}

dbvn <- function(v, mu = c(0, 0), Sigma = matrix(c(1, 0.6, 0.6, 1), 2, 2)) {
  x <- matrix(v - mu, ncol = 1)
  Sinv <- solve(Sigma)
  1 / (2 * pi * sqrt(det(Sigma))) * exp(-0.5 * as.numeric(t(x) %*% Sinv %*% x))
}

hdr_1d <- ggplot() +
  geom_pdf(fun = bimodal, xlim = c(-5, 5), hdr_xlim = c(-7, 7), shade_hdr = 0.8) +
  labs(x = "x", y = "density")

hdr_2d <- ggplot() +
  geom_pdf_2d(fun = dbvn, xlim = c(-3, 3), ylim = c(-3, 3), probs = c(0.5, 0.8, 0.95), n = 60) +
  coord_equal() +
  labs(x = "x", y = "y")

hdr_1d + hdr_2d
```

The bivariate PMF layer uses the same vector-argument convention as the
bivariate PDF layer, but evaluates on a lattice. Figure
\@ref(fig:pmf-2d-workflow) shows the same independent product-binomial mass
function in point and tile modes. Point mode maps probability to area, while
tile mode maps probability to fill.

```{r pmf-2d-workflow, fig.width=7, fig.height=3, out.width="95%", fig.cap="A bivariate probability mass function displayed as a balloon plot and as a tile heatmap."}
dbinom2 <- function(v, sizes = c(10, 10), probs = c(0.35, 0.65)) {
  dbinom(v[1], sizes[1], probs[1]) * dbinom(v[2], sizes[2], probs[2])
}

pmf2_point <- ggplot() +
  geom_pmf_2d(fun = dbinom2, xlim = c(0, 10), ylim = c(0, 10)) +
  scale_size_area(max_size = 7) +
  coord_equal() +
  labs(title = "point", x = "x", y = "y", size = "mass")

pmf2_tile <- ggplot() +
  geom_pmf_2d(fun = dbinom2, xlim = c(0, 10), ylim = c(0, 10), type = "tile") +
  coord_equal() +
  labs(title = "tile", x = "x", y = "y", fill = "mass")

pmf2_point + pmf2_tile
```

Discrete distributions follow the same ideas on a finite or integer support.
The PMF layer uses lollipops, discrete CDF/quantile/survival layers use step
functions, and discrete HDRs include all support points tied at an HDR cutoff.
Figure \@ref(fig:discrete-support) uses a deliberately nonconsecutive support
to show why the support itself is part of the statistical object: cumulative
sums and inverse lookups must be performed on the declared support, not on all
integers between the minimum and maximum. For examples involving ties and
bivariate mass functions, see the probability and numerical-conversion
vignettes.

```{r discrete-support, fig.width=7, fig.height=3, out.width="95%", fig.cap="A discrete distribution on a nonconsecutive support. Both layers resolve the same lower cumulative shading request; the discrete CDF is derived from the same PMF by cumulative summation over the declared support."}
disc_support <- c(0, 1, 3, 4, 8)
disc_mass <- c(0.12, 0.28, 0.20, 0.25, 0.15)
disc_pmf <- function(x) {
  out <- numeric(length(x))
  idx <- match(x, disc_support)
  out[!is.na(idx)] <- disc_mass[idx[!is.na(idx)]]
  out
}

pmf_plot <- ggplot() +
  geom_pmf(fun = disc_pmf, support = disc_support, p = 0.8) +
  labs(title = "PMF", x = "support", y = "mass")

cdf_plot <- ggplot() +
  geom_cdf_discrete(pmf_fun = disc_pmf, support = disc_support, p = 0.8) +
  labs(title = "CDF from PMF", x = "support", y = "cumulative mass")

pmf_plot + cdf_plot
```

## Empirical distributions and censored summaries

When users have data, the empirical workflow compares sample summaries with distributional structure.
For complete samples, `geom_ecdf()` and `geom_eqf()` draw empirical
distribution and quantile functions. Their confidence bands use the
Dvoretzky--Kiefer--Wolfowitz inequality with Massart's sharp constant
[@massart1990tight]. For a fully specified distribution $F$, the ECDF band
\[
  \hat F_n(x) \pm
  \sqrt{\frac{\log(2/\alpha)}{2n}}
\]
has simultaneous coverage at least $1-\alpha$, clipped to $[0,1]$. Grouped
ECDFs receive group-specific bands, as in Figure \@ref(fig:ecdf-workflow).
The empirical quantile band is obtained by inverting these CDF limits. This
inversion is visually useful but produces wide or partially uninformative tail
regions when $p\pm\varepsilon$ reaches 0 or 1, which is a property of the DKW
bound rather than a plotting artifact.

The band drawn in Figure \@ref(fig:ecdf-workflow) is the direct finite-sample
DKW/Massart band. The package vignette `coverage-simulation` gives additional
simulation checks across distributions, sample sizes, and replicate counts.

```{r ecdf-workflow, fig.width=6, fig.height=3, fig.cap="Grouped ECDFs with DKW/Massart confidence bands. Each group receives a separate band computed from its own sample size."}
set.seed(3101)
ecdf_data <- data.frame(
  x = c(rnorm(80, 0, 1), rnorm(120, 1.2, 1)),
  group = rep(c("A", "B"), c(80, 120))
)

ggplot(ecdf_data, aes(x = x, colour = group, fill = group)) +
  geom_ecdf() +
  labs(x = "x", y = "empirical CDF", colour = "group", fill = "group")
```

The same complete-data sample can be viewed on the cumulative-hazard scale.
This is useful when the theoretical hazard has a simpler visual form than the
CDF. For exponential data, the cumulative hazard is linear, so departures from
linearity are easier to see after transformation. The complete-data
`geom_echf()` layer applies the monotone transform
\(H(x)=-\log\{1-F(x)\}\) to the ECDF and to the DKW limits. Where the upper DKW
limit reaches 1, the transformed upper limit is infinite; the figure clips that
edge for display. Figure \@ref(fig:echf-complete) shows the same sample on the
ECDF and empirical cumulative-hazard scales.

```{r echf-complete, fig.width=7, fig.height=3, out.width="95%", fig.cap="The same complete exponential sample shown as an ECDF and as an empirical cumulative hazard. The ECDF is curved under an exponential model, while the cumulative-hazard target is linear. The displayed ECHF upper band is clipped where the transformed upper DKW bound saturates."}
set.seed(3104)
haz_data <- data.frame(x = rexp(100, rate = 0.6))
haz_xlim <- c(0, max(haz_data$x))

cdf_complete <- ggplot(haz_data, aes(x = x)) +
  geom_ecdf() +
  geom_cdf(fun = function(x) pexp(x, rate = 0.6), xlim = haz_xlim, color = "#D55E00") +
  labs(title = "ECDF", x = "x", y = "distribution function")

echf_complete <- ggplot(haz_data, aes(x = x)) +
  geom_echf(band_max = 6) +
  geom_chf(fun = function(x) 0.6 * x, xlim = haz_xlim, color = "#D55E00") +
  coord_cartesian(ylim = c(0, 6)) +
  labs(title = "ECHF", x = "x", y = "cumulative hazard")

cdf_complete + echf_complete
```

Diagnostic layers use the same probability-scale argument. `geom_ppplot()`
compares theoretical and empirical probabilities, `geom_qqplot()` compares
theoretical and sample quantiles, and `geom_spplot()` applies Michael's
arcsine-square-root stabilization [@michael1983stabilized] to the PP plot.
Observations enter through the non-positional `sample` aesthetic, as in
`ggplot2::geom_qq()`: \CRANpkg{ggplot2} transforms position aesthetics before
a layer's statistic runs, so a positional input would hand the null
distribution transformed values whenever an axis carries a transformation,
whereas `sample` guarantees that the null CDF or quantile function always
receives the raw data. Because the PP and SP bands are continuous-null
procedures, requesting one requires declaring `null_type = "continuous"`.
When the null distribution is fully specified via an input function `fun`, the DKW/Massart band is a
finite-sample simultaneous band for the probability integral transform. When
parameters are estimated from the same data being plotted, that argument no
longer gives the stated finite-sample simultaneous calibration. This distinction
is important in practice. In teaching, it is common to compare a sample with
`pnorm` after estimating mean and standard deviation from the same data.
\pkg{ggfunction} will draw the requested visual band, but it does not perform a
Lilliefors [@lilliefors1967] or other composite-null adjustment. The plotted
band can still be a useful exploratory diagnostic; users who need a calibrated
composite-null goodness-of-fit procedure should use a method designed for that
setting.

```{r diagnostic-workflow, fig.width=7, fig.height=4.6, out.width="95%", fig.cap="QQ and stabilized probability plots for one sample against a fully specified standard normal null. Fitted-null diagnostics use the same visual band but require separate calibration for formal testing."}
set.seed(3102)
diag_data <- data.frame(x = rt(90, df = 4) / sqrt(2))

qq_plot <- ggplot(diag_data, aes(sample = x)) +
  geom_qqplot(fun = qnorm, colour = "black", size = 1.3) +
  coord_equal() +
  labs(title = "QQ", x = "theoretical quantile", y = "sample quantile")

sp_plot <- ggplot(diag_data, aes(sample = x)) +
  geom_spplot(fun = pnorm, null_type = "continuous", colour = "black", size = 1.3) +
  coord_equal() +
  labs(title = "SP", x = "theoretical", y = "observed")

qq_plot + sp_plot
```

For right-censored data, `geom_ecdf_km()` computes the Kaplan--Meier
product-limit estimator [@kaplan1958nonparametric] from observed times and an
event indicator. It uses risk sets and displays censoring marks by default. Its
confidence band uses Greenwood's variance estimator [@greenwood1926natural] and
the equal-precision construction of Nair [-@nair1984confidence], which is
simultaneous and asymptotic. The companion `geom_echf_na()` layer computes
the Nelson--Aalen cumulative hazard estimator [@nelson1972theory;
@aalen1978nonparametric] and uses the implemented pointwise normal confidence
intervals. Figure \@ref(fig:survival-workflow) shows the two estimators side
by side.
For censored observations, ECDF-based complete-data summaries are no longer
appropriate; the risk-set calculations in the Kaplan--Meier and Nelson--Aalen
estimators are essential.

At each distinct event time $t_j$, let $d_j$ denote the number of events and
$n_j$ the number at risk immediately before $t_j$. The Kaplan--Meier estimator
displayed by `geom_ecdf_km()` is
\[
  \hat S(t) = \prod_{t_j \le t} \left(1-\frac{d_j}{n_j}\right),
\]
and the Nelson--Aalen estimator displayed by `geom_echf_na()` is
\[
  \hat H(t) = \sum_{t_j \le t} \frac{d_j}{n_j}.
\]
The package computes both from the same internal risk-set table. This shared
tabulation keeps event counts, censoring counts, risk sets, and variance
estimates consistent between the two censored-data layers.

The confidence displays intentionally differ. The Kaplan--Meier band uses
Greenwood's variance estimate and an equal-precision critical value following
Nair's construction. The displayed limits are additive on the survival scale,
\(\hat S(t) \pm c_{\mathrm{EP}}\mathrm{se}\{\hat S(t)\}\), and are clipped to
\([0,1]\). The critical value is computed from Greenwood-scale endpoints at the
first and last valid event times, so the band is a simultaneous large-sample
band over the displayed event-time range, not an exact finite-sample band.
These default endpoints make the display a data-adaptive plug-in version of
Nair's procedure; a prespecified equal-precision domain can be supplied
through `ep_range = c(a_L, a_U)`, from which the critical value is then
computed. Event times where Greenwood's variance is undefined---notably a
terminal event at which all remaining subjects fail---are excluded from the
band rather than assigned a degenerate zero-width interval, and the last
defined interval extends flat to the end of observed follow-up as a geometric
anchor rather than a new interval. The point estimates likewise extend
through trailing censoring, and all-censored groups draw $\hat S \equiv 1$
over their observation window. The
Nelson--Aalen intervals use the Nelson variance estimate
\[
  \widehat{\mathrm{Var}}\{\hat H(t)\} =
  \sum_{t_j \le t} \frac{d_j}{n_j^2},
\]
and a normal critical value at each time point. Its intervals are symmetric on
the cumulative-hazard scale before the lower limit is clipped at zero. It is
therefore pointwise, not simultaneous. The visual encodings intentionally
differ: the simultaneous Kaplan--Meier band is drawn as a ribbon, while the
pointwise Nelson--Aalen intervals are drawn as capped error bars at event
times. The examples use this distinction deliberately.

```{r survival-workflow, fig.width=7, fig.height=4.6, out.width="95%", fig.cap="Censored-data summaries from the same simulated sample. The smooth orange curves show the generating Exp(0.45) survival and cumulative-hazard targets. Left: Kaplan--Meier survival with censor marks and a simultaneous Greenwood/Nair band. Right: Nelson--Aalen cumulative hazard with pointwise normal intervals."}
set.seed(3103)
n_surv <- 90
event_time <- rexp(n_surv, rate = 0.45)
censor_time <- rexp(n_surv, rate = 0.25)
surv_data <- data.frame(
  time = pmin(event_time, censor_time),
  status = as.integer(event_time <= censor_time)
)

km_plot <- ggplot(surv_data, aes(x = time, status = status)) +
  geom_ecdf_km(censor_size = 1.5) +
  geom_survival(fun = function(x) exp(-0.45 * x), xlim = c(0, max(surv_data$time)),
                color = "#D55E00", linewidth = 0.5) +
  labs(title = "Kaplan-Meier", x = "time", y = "survival")

na_plot <- ggplot(surv_data, aes(x = time, status = status)) +
  geom_echf_na() +
  geom_chf(fun = function(x) 0.45 * x, xlim = c(0, max(surv_data$time)),
           color = "#D55E00", linewidth = 0.5) +
  labs(title = "Nelson-Aalen", x = "time", y = "cumulative hazard")

km_plot + na_plot
```

# Implementation, numerical behavior, and performance

## ggplot2 stat-geom architecture

Each user-facing layer constructs one or more \CRANpkg{ggplot2} layers. The
stat component validates inputs, evaluates functions or tabulates empirical
summaries, and returns a data frame with computed variables such as
`x`, `y`, `ymin`, `ymax`, or `probs`. The geom
component draws those data using existing \CRANpkg{ggplot2} infrastructure when
possible: paths and ribbons for curves and bands, rasters and contours for
scalar fields, and step-like geoms for discrete and empirical functions.

The two main delegated components are explicit. Vector and stream fields are
wrappers around \CRANpkg{ggvfields} vector-field and stream-field layers. Bivariate
PDF HDR displays are thin wrappers around \CRANpkg{ggdensity}'s functional HDR
geoms; the raster mode for `geom_pdf_2d()` uses the scalar-field machinery
instead. This delegation keeps \pkg{ggfunction} focused on function interfaces,
probability semantics, and validation, while not duplicating other work maintained by the authors.

The layer constructors also provide a predictable data contract. Most users do
not inspect the built data, but doing so is useful for tests, extensions, and
debugging. Table \@ref(tab:data-contract) summarizes representative computed
variables. These are not new aesthetics that users must supply; they are values
created by the stat and then consumed by the geom.

```{r data-contract}
#| tab.cap: "Representative computed variables returned by ggfunction stats."
data_contract <- data.frame(
  Family = c(
    "curves",
    "fields",
    "continuous prob.",
    "discrete prob.",
    "empirical",
    "censored",
    "diagnostics"
  ),
  Variables = c(
    "\\code{x}, \\code{y}",
    "\\code{x}, \\code{y}, \\code{z}",
    "\\code{x}, \\code{y}, shade vars",
    "\\code{x}, \\code{y}, mass/HDR",
    "\\code{x}, \\code{y}, band limits",
    "\\code{x}, \\code{y}, band/interval/censor",
    "\\code{x}, \\code{y}, order/probs"
  ),
  Geom = c(
    "path/area",
    "raster/contour",
    "area/path/ribbon",
    "lollipop/step",
    "step/ribbon",
    "step/ribbon/bars/marks",
    "points/ribbon/line"
  ),
  check.names = FALSE
)
knitr::kable(data_contract, escape = FALSE)
```

This contract is also how the package is tested. Unit tests build plots with
`ggplot_build()`, inspect the computed data frames, and compare
band limits, support handling, route accuracy, and error messages with expected
values. The tests therefore exercise the same data that a downstream extension
would see.

## Numerical conversions

The conversion code is intentionally conservative. Native functions are used
directly. Derived functions are tagged internally as approximate, and their
accuracy depends on smoothness, support, tail behavior, grid resolution, and
whether the supplied function is numerically stable. The main conversion methods
are:

- PDF to CDF: `stats::integrate()` over the declared support.
- CDF to PDF: central finite differences with support-aware endpoint handling,
  implemented directly rather than through an external numerical-differentiation
  dependency.
- CDF to quantile: bracketing and `stats::uniroot()`.
- Quantile to CDF: `stats::approxfun()` interpolation on a probability grid.
- Survival to CDF or cumulative hazard: algebraic transformations.
- Survival or PDF to hazard: the logarithmic derivative $-\tfrac{d}{dx}\log
  S(x)$, or $f$ divided by an upper-tail integral of $f$; both remain stable
  deep in the tail, where forming $f/(1-F)$ loses all precision once $F$
  rounds to 1.
- Hazard to cumulative hazard: `stats::integrate()` from the hazard origin,
  with exact endpoint values ($H = \infty$, hence $F = 1$ and $S = 0$, at the
  upper support endpoint) rather than attempted divergent integrals.
- Hazard to PDF: $f(x) = h(x)\exp\{-H(x)\}$ after deriving $H$.
- Discrete conversions: cumulative sums over sorted support and generalized
  inverse lookup.

These routes are meant to support plotting, diagnostics, and teaching examples,
not to replace distribution-specific numerical libraries. The numerical CDF
derived from a smooth bounded density can be extremely accurate on a moderate
grid. The same default settings may be inadequate for a density with sharp
spikes near a boundary or for a heavy-tailed distribution whose visually
interesting mass lies outside the displayed window. The API therefore exposes
the most important controls--the declared support, the display window, the grid
resolution, and the hazard integration origin--instead of hiding all numerical
choices behind a single plotting call.

For hazards, the lower integration endpoint deserves special mention. If the
support has a finite lower endpoint and `hf_lower` is left at its default of
`-Inf`, the package uses the lower support endpoint. This makes common
nonnegative lifetime distributions behave naturally when users write
`support = c(0, Inf)`. If neither the support nor the hazard origin is finite,
integrating a hazard can be ill-posed; the layer then warns or fails rather
than silently drawing a misleading curve.

Tables \@ref(tab:technical-accuracy) and \@ref(tab:technical-performance)
report representative local validation results from the accuracy and benchmark
scripts in `inst/benchmarks/`. These scripts are manual validation
aids rather than CRAN tests. They install the local package, compare fixed-grid
function values, or time `ggplot_build()` for representative plots. The saved
outputs used here are in `inst/benchmarks/results/` and record the run date,
platform, \proglang{R} version, package versions, grid details, and benchmark
iteration counts.

```{r validation-results-setup, include=FALSE}
find_result_file <- function(prefix) {
  candidates <- c(
    file.path("..", "inst", "benchmarks", "results"),
    file.path("inst", "benchmarks", "results")
  )
  for (dir in candidates) {
    files <- list.files(dir, pattern = paste0("^", prefix, ".*\\.csv$"),
                        full.names = TRUE)
    if (length(files) > 0L) {
      return(sort(files, decreasing = TRUE)[1L])
    }
  }
  stop("No saved benchmark result file found for prefix: ", prefix, call. = FALSE)
}

accuracy_results <- utils::read.csv(find_result_file("accuracy-"), check.names = FALSE)
benchmark_results <- utils::read.csv(find_result_file("benchmark-"), check.names = FALSE)
validation_context <- accuracy_results[1L, c(
  "run_time", "platform", "r_version", "ggfunction_version",
  "ggplot2_version", "ggdensity_version", "ggvfields_version",
  "bench_version"
)]
```

The reported run (`r validation_context$run_time`) was on
`r validation_context$platform` using
\proglang{R} `r sub("^R version ", "", validation_context$r_version)`,
\pkg{ggfunction} `r validation_context$ggfunction_version`,
\CRANpkg{ggplot2} `r validation_context$ggplot2_version`,
\CRANpkg{ggdensity} `r validation_context$ggdensity_version`, and
\CRANpkg{ggvfields} `r validation_context$ggvfields_version`. The benchmark
script used \CRANpkg{bench} [@bench] `r validation_context$bench_version`. Accuracy
grids used 401 x-values for the normal, exponential, Weibull, and beta checks,
199 probabilities for quantile checks, and the 21-point binomial support.

```{r technical-accuracy}
#| tab.cap: "Representative numerical conversion accuracy from saved validation output. Errors are maximum absolute differences on the fixed grids described in the text."
format_route <- function(x) {
  vapply(strsplit(x, "->", fixed = TRUE), function(parts) {
    paste(toupper(parts), collapse = " $\\to$ ")
  }, character(1))
}
accuracy_table <- data.frame(
  Route = format_route(accuracy_results$route),
  Distribution = accuracy_results$distribution,
  Grid = accuracy_results$n_grid,
  "Max error" = formatC(accuracy_results$max_abs_error, format = "e", digits = 2),
  Note = ifelse(nchar(accuracy_results$notes) == 0, "--", accuracy_results$notes),
  check.names = FALSE
)
knitr::kable(accuracy_table, escape = FALSE)
```

```{r technical-performance}
#| tab.cap: "Representative local performance checks from saved benchmark output. Timings are medians over 10 iterations of plot construction, not full device rendering."
performance_table <- data.frame(
  Case = benchmark_results$case,
  Domain = benchmark_results$grid_domain,
  Median = trimws(benchmark_results$median),
  Iterations = benchmark_results$n_itr,
  check.names = FALSE
)
knitr::kable(performance_table)
```

The tables should be read as reproducibility checks, not universal guarantees.
Heavy tails, discontinuities, bounded supports, singular hazards, and nearly
flat CDF regions are harder cases. Users can improve behavior by supplying the
native distribution function, declaring finite `support`, increasing
`n`, or setting a finite `hf_lower` for hazards whose origin is known.

## Performance and scaling

For native one-dimensional functions, build time is dominated by evaluating the
function at `n` grid points and by \CRANpkg{ggplot2} layer construction, so
the cost scales roughly linearly in `n`. Scalar fields and bivariate
density displays evaluate an $n_x \times n_y$ grid, so their computational cost
scales like $n^2$. Conversions involving integration or root-finding are more
expensive than native evaluations because they call numerical routines at many
grid points. Stream fields depend additionally on ODE settings and the behavior
of the vector field; rendering time can dominate for dense contours, HDRs, and
streamlines.

## Validation and warnings

The package validates common failure modes before plotting, in two tiers.
Structural invalidity---a source function returning the wrong type or
length, or non-finite or negative mass---always errors, because no coherent
display exists for such input. Distributional plausibility is a diagnostic
tier: densities are checked for approximate normalization over their
declared support, PMFs for approximate unit mass, cumulative sources for
monotonicity and the unit interval, computed survival, quantile, hazard, and
cumulative-hazard curves for their defining shape constraints, and HDR
computation domains for omitted probability mass. These diagnostics alert
without aborting---the declared object is presumed to be what the layer name
says it is---and can be disabled per layer via `check = FALSE` or globally
via the package option. Separately, a request for a quantity the declared
object does not possess warns unconditionally rather than fabricating an
answer: a shading probability exceeding the mass attained on a truncated
support warns and shades the attainable region, and a discrete quantile
function simply ends at the attained mass. Invalid source combinations are
rejected. Most distribution geoms require
exactly one source function, but hazard layers also accept the documented
bundle `pdf_fun` + `cdf_fun`, which avoids deriving either component
numerically.

The validation policy aims to catch mistakes that are easy to make in plotting
code. Examples include using the displayed window as if it were the full
support, supplying an unnamed `args` list, drawing a density that integrates far
from one, evaluating a CDF-derived PDF outside the support, or asking for a
quantile route without finite bracketing information. When the package can
continue safely, it warns and returns missing values for failed points. When
the requested route is logically ambiguous, it errors before building the plot;
for example, `geom_pdf(cdf_fun = pnorm, survival_fun = function(x) 1 - pnorm(x))`
supplies two possible sources for the same PDF target and is rejected.

The checks are deliberately lightweight enough for interactive plotting. They
do not prove that an arbitrary user-supplied function is a valid distribution
function. In particular, they cannot rescue a discontinuous CDF from finite
differencing artifacts, identify every singular hazard, or choose a universally
appropriate HDR grid. The documentation and numerical-conversion vignette
therefore emphasize native distribution functions and explicit support
declarations whenever accuracy matters.

The package also includes supporting infrastructure for submission and
maintenance. Documentation generated with \CRANpkg{roxygen2} [@roxygen2]
records the accepted source combinations, approximation routes, support/window
distinction, and warning behavior for each user-facing layer. Vignettes move
longer teaching examples and validation
stories out of the article: probability geoms, empirical geoms, numerical
conversions, survival analysis, and DKW coverage simulation each have a focused
home. Manual benchmark scripts in `inst/benchmarks/` provide reproducible
accuracy and timing checks without running during `R CMD check`. This separation
keeps CRAN checks fast while preserving evidence for manuscript claims.

# Composability and limitations

Because \pkg{ggfunction} returns standard \CRANpkg{ggplot2} layers, its outputs
compose with scales, coordinates, themes, facets, and annotations in ways \CRANpkg{ggplot2} users have come to expect. A theoretical
curve can be overlaid on an ECDF, a survival curve can be added to a
Kaplan--Meier estimate, and a scalar field can be recolored with ordinary fill
scales. This composability is the main practical benefit of implementing
functions inside the \CRANpkg{ggplot2} layer system.

Composability also makes the package useful in documents and teaching material.
A probability layer can sit next to a simulation layer in a \CRANpkg{patchwork}
layout [@patchwork], use the same theme as the surrounding analysis, or inherit
a coordinate system that zooms the display without changing the distributional
calculation.
The support/window distinction is especially important here: a plot can compute
probability content over the full support while using `coord_cartesian()` to
focus on the region where the visual comparison matters.

The same design also defines the limits. Functions are layer parameters, not
aesthetics, so faceting over a data-frame column of function objects is not
supported in the ordinary grammar-of-graphics sense. Numerical conversions are
approximations and can fail for difficult distributions. HDRs are grid-based,
so their thresholds depend on the computational domain and grid resolution.
Two-dimensional grids can become expensive at high resolution. Censored-data
confidence displays have the validity described in the implementation:
Kaplan--Meier uses the implemented simultaneous EP Greenwood/Nair band, while
Nelson--Aalen uses implemented pointwise normal intervals.

There are also statistical limits that no plotting interface can remove.
Probability shading is only as meaningful as the distribution supplied to the
layer. A diagnostic band for a fully specified null model has a different
interpretation from the same band drawn after fitting parameters. A gridded HDR
computed on too narrow a domain may have the wrong threshold. A survival curve
with heavy censoring in the tail can have a visually wide confidence display
because the risk set is small. The package tries to make these situations
visible, but the user remains responsible for choosing a model and interpreting
the plot in context.

# Related work

\pkg{ggfunction} is closest to tools that bridge statistical or mathematical
objects and graphics. The compact summaries below describe the closest
\CRANpkg{ggplot2} extensions and adjacent packages. Table
\@ref(tab:related-core) emphasizes function-layer design; Table
\@ref(tab:related-adjacent) emphasizes empirical, diagnostic, and
survival displays. The intent is complementarity rather than replacement:
several packages solve different problems better than \pkg{ggfunction} does.

```{r related-core}
related_core <- data.frame(
  Package = c(
    "\\code{ggplot2::stat\\_function()}",
    "\\CRANpkg{ggvfields}",
    "\\CRANpkg{metR}",
    "\\CRANpkg{ggformula}/\\CRANpkg{mosaicCalc}"
  ),
  Relationship = c(
    "native univariate function-curve baseline",
    "delegated vector-field and stream-field rendering",
    "stronger support for precomputed gridded fields",
    "formula-oriented front ends for calculus and teaching"
  ),
  check.names = FALSE
)
knitr::kable(
  related_core,
  escape = FALSE,
  longtable = TRUE,
  caption = "Closest function-layer and field-display packages."
)
```

```{r related-adjacent}
related_adjacent <- data.frame(
  Package = c(
    "\\CRANpkg{ggdist}",
    "\\CRANpkg{ggdensity}",
    "\\CRANpkg{hdrcde}",
    "\\CRANpkg{qqplotr}",
    "\\CRANpkg{survminer}/\\CRANpkg{ggsurvfit}"
  ),
  Relationship = c(
    "broader distribution displays",
    "bivariate HDR engine used here",
    "HDR objects outside layer taxonomy",
    "wider band/detrend menu",
    "risk-table/test workflows"
  ),
  check.names = FALSE
)
knitr::kable(
  related_adjacent,
  escape = FALSE,
  longtable = TRUE,
  caption = "Adjacent distribution, diagnostic, and survival-display packages."
)
```

\CRANpkg{ggplot2} [@wickham2010layered; @wickham2016ggplot2] and Wilkinson's
grammar [@wilkinson2005grammar] provide the conceptual and software foundation.
\CRANpkg{ggdist} [@kay2024ggdist] is broader for uncertainty visualization.
\CRANpkg{ggdensity} [@otto2023ggdensity] supplies the bivariate HDR machinery
used here. \CRANpkg{ggvfields} [@ggvfields] supplies vector-field and
stream-field infrastructure. \CRANpkg{metR} [@metR] and \CRANpkg{hdrcde}
[@hdrcde] cover adjacent meteorological and HDR-focused use cases.
\CRANpkg{ggformula} [@ggformula], \CRANpkg{mosaic} [@pruim2017mosaic], and
\CRANpkg{mosaicCalc} [@mosaicCalc] provide formula-oriented front ends for
functions, distributions, calculus, and introductory statistics; \pkg{ggfunction}
instead emphasizes native stats/geoms, support-aware probability semantics, and
conversion routes among distribution representations.

The package boundaries are therefore partly philosophical. \CRANpkg{ggdist}
starts from distributions and uncertainty summaries and provides a rich visual
vocabulary for communicating posterior draws, intervals, slabs, and dots.
\pkg{ggfunction} starts from ordinary \proglang{R} functions and asks how they
should be evaluated inside a \CRANpkg{ggplot2} layer. \CRANpkg{ggdensity}
focuses on bivariate density estimation and probability-calibrated contours;
\pkg{ggfunction} uses that machinery when the user supplies a theoretical
bivariate density. \CRANpkg{ggvfields} focuses on vector-field displays and
integration; \pkg{ggfunction} exposes those displays through the dimensional
taxonomy so scalar and vector-valued functions share a naming scheme.

For diagnostic plots, \CRANpkg{qqplotr} [@almeida2018qqplotr; @qqplotr] is the
closest \CRANpkg{ggplot2} extension, providing Q-Q and P-P points, reference
lines, confidence-band constructions, and detrended displays. \pkg{ggfunction}
overlaps through `geom_qqplot()` and `geom_ppplot()`, and adds `geom_spplot()`
within the same support-aware probability framework used by its ECDF and
quantile layers; users who need a wider diagnostic-band or detrending menu
should use \CRANpkg{qqplotr}. For right-censored data, \CRANpkg{survminer}
[@survminer] and \CRANpkg{ggsurvfit} [@ggsurvfit] provide publication-oriented
time-to-event figures with survival-analysis furniture such as risk tables,
censoring summaries, stratified comparisons, and ready-to-share themes;
\CRANpkg{ggsurvfit} is itself modular and \CRANpkg{ggplot2}-compatible. The
censored-data layers in \pkg{ggfunction} emphasize a lower-level role inside
the same support-aware probability framework used by its other empirical geoms:
`geom_ecdf_km()` and `geom_echf_na()` share one risk-set tabulation, use
deliberately different confidence encodings for simultaneous and pointwise
displays, and overlay theoretical survival or cumulative-hazard curves as
ordinary layers. They do not aim to replace survival-workflow packages.

This scope keeps the package lightweight. It does not define distribution
objects, fit models, estimate bivariate densities from samples, or implement a
general symbolic mathematics system. Instead it provides a bridge from function
definitions and empirical summaries to ordinary \CRANpkg{ggplot2} layers. That
bridge is useful precisely because it leaves the rest of the graphics system
unchanged.

# Discussion and future directions

While \pkg{ggfunction} is a large package, it makes a narrow commitment: take a mathematical or statistical
function, evaluate it safely enough for plotting, and return a standard
\CRANpkg{ggplot2} layer. That commitment covers common teaching and diagnostic
tasks without asking users to build grids, integrate densities, or hand-code
confidence ribbons, and in doing so alleviates pain points experienced by
countless \CRANpkg{ggplot2} users in a thoughtful way.

Future work could add better support for collections of functions, richer
interfaces for distribution objects, more user controls for numerical
conversion tolerances, conditional distributions and HDRs, additional
censored-data confidence displays, and higher-level wrappers for common
teaching workflows.
The benchmark and accuracy scripts added during package hardening provide a
starting point for tracking these changes over time.

The most natural next step is support for function collections. Many classroom
and applied examples compare a family of functions--normal densities with
different scales, hazard functions under several parameter settings, or
solution curves from different initial conditions. Today these can be overlaid
by adding multiple layers manually. A future interface could make such
collections more data-like while respecting the fact that functions are not
ordinary aesthetics.

Another direction is more explicit numerical control. The defaults are chosen
for readable interactive graphics, but some users will want tighter integration
tolerances, adaptive grids, or route-specific diagnostics. Exposing those
controls without turning simple plots into numerical-analysis exercises is a
design challenge. The validation scripts introduced here give the package a
baseline against which such changes can be checked.

# Acknowledgments

James Otto contributed early drafts of the CDF and ECDF-related stats and
geoms. Rodney Sturdivant contributed early design feedback. The authors used
OpenAI's GPT 5.5 and Anthropic's Claude Opus 4.8 extensively as programming and
drafting assistants during package development and manuscript revision; the
authors reviewed and are responsible for the package code, examples, and
manuscript claims.
